\documentclass{my_memo}

\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{longtable}
\sloppy

\lstset{
  language=Fortran,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue!60!black},
  commentstyle=\color{green!40!black},
  stringstyle=\color{red!60!black},
  columns=fullflexible,
  breaklines=true,
  frame=single,
  framesep=4pt,
  showstringspaces=false,
  numbers=none
}
\lstdefinestyle{sh}{language=bash,keywordstyle=\color{black},basicstyle=\ttfamily\small}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\um}{\ensuremath{\,\mu\mathrm{m}}}
\newcommand{\Cabs}{\ensuremath{C_{\rm abs}}}
\newcommand{\Cpol}{\ensuremath{C_{\rm pol}}}
\newcommand{\falign}{\ensuremath{f_{\rm align}}}

\title{Polarization in SEDust: Implementation Notes}
\author{Kwang-il Seon}
\date{2026 July 20}

\begin{document}
\maketitle

\begin{abstract}
\noindent
These are maintenance notes for the polarization capability added to SEDust:
where the polarized optics come from, how they are assembled, which
implementation choices were made and why, and how far the numbers have been
checked. The reader is assumed to be someone extending or repairing this code,
not someone calling it. The calling interface is documented in
\code{SEDust\_user\_manual.pdf}; the underlying physics is documented in
\code{aligned\_grain\_polarization.pdf}. Neither is repeated here.
\end{abstract}

\tableofcontents

%====================================================================
\section{Scope, and what lives elsewhere}

Three documents cover this work and they do not overlap
(Table~\ref{tab:whichdoc}). This one answers only the question \emph{how was
this built and why that way}. Where physics is needed to justify a coding
choice, the relevant section of the theory report is cited rather than
restated.

\begin{table}[h]
\centering
\begin{tabular}{@{}p{0.46\textwidth}p{0.46\textwidth}@{}}
\toprule
Question & Document \\
\midrule
Why polarization at all; alignment theory; the polarized transfer equation;
sign and index conventions; validation targets &
\code{aligned\_grain\_polarization.pdf} (\S1--\S5, \S9--\S11) \\[2pt]
How an RT code calls the library: signatures, units, what is already
integrated &
\code{SEDust\_user\_manual.pdf}, \S5 \\[2pt]
How the code is built, why each choice was made, what was measured &
this note \\
\bottomrule
\end{tabular}
\caption{Division of the polarization documentation.}
\label{tab:whichdoc}
\end{table}

\paragraph{Where this is going.} The polarization capability is not an end in
itself. The destination is the polarized radiative transfer of
Seon (2018): dust scattering plus dichroic extinction by aligned grains,
computed with grain optics derived from first principles, in the Monte Carlo
formalism of Peest et al.\ (2017) as extended to aligned spheroids by
Peest et al.\ (2023). Under that contract MoCafe supplies the alignment
assumptions (the \falign\ profile and the local magnetic-field geometry) and
SEDust returns every material quantity in matrix form: the $4\times4$
extinction matrix $K(\theta_i)$ with dichroism and birefringence, the phase
matrix $Z_{\rm al}(\theta_i;\theta_s,\phi)$ of the aligned population, the
random-orientation remainder, and the scattering cross sections. Dichroic
extinction, polarized emission, birefringence, and the orientation-resolved
cross-section table cover the extinction and emission side of that contract; the
scattering side --- the fixed-orientation Mueller matrix, the extinction matrix
built from forward amplitudes, and their library interface --- is now built and
verified, and is documented in \S\ref{sec:aligned}. What remains is MoCafe's
consumption: the frame rotations, direction sampling, peel-off, and the
$\exp(-K\tau)$ transfer step, none of which is a SEDust responsibility.

%====================================================================
\section{The data source}\label{sec:source}

Everything polarized derives from one file, the orientation-resolved
Draine \& Hensley (2021) astrodust spheroid table:
\begin{lstlisting}[style=sh]
data/dielectric/q_DH21Ad_P0.20_Fe0.00_1.400.dat.gz
\end{lstlisting}
(porosity $P=0.20$, iron fraction $f_{\rm Fe}=0$, axial ratio $b/a=1.400$).
Its layout is not self-describing, so it is recorded here
(Table~\ref{tab:filelayout}).

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
\toprule
Property & Value \\
\midrule
header                & 12 lines, skipped \\
payload order         & \code{((Q(jw,jr,jori), jw=0,1128), jr=0,168), jori=1,3} \\
repeated              & three times: $Q_{\rm ext}$, then $Q_{\rm abs}$, then $Q_{\rm sca}$ \\
one record on disk    & the 169 sizes of one $(jw,\,jori)$ pair \\
records read          & $3~\text{quantities}\times3~\text{orientations}\times1129$ \\
total values          & $3\times3\times1129\times169 = 1{,}717{,}209$ \\
wavelength axis       & \code{data/dielectric/DH21\_wave}, 1129 nodes \\
size axis             & \code{data/dielectric/DH21\_aeff}, 169 nodes \\
axis file format      & two title lines, then the values free-format \\
\bottomrule
\end{tabular}
\caption{Layout of the orientation-resolved $Q$ table. The wavelength and size
axes are \emph{not} parsed out of the header; they come from the companion grid
files, which list the nodes the table was computed on.}
\label{tab:filelayout}
\end{table}

\paragraph{Orientation index.} The convention is the trap most likely to bite a
future reader, and it is worth stating in full (Table~\ref{tab:jori}); see also
\S3.6 of the theory report.

\begin{table}[h]
\centering
\begin{tabular}{@{}cll@{}}
\toprule
\code{jori} & geometry & meaning \\
\midrule
1 & $\mathbf{k}\parallel\hat{a}$        & propagation along the symmetry axis \\
2 & $\mathbf{k}\perp\hat{a}$, $\mathbf{E}\parallel\hat{a}$ & field along the symmetry axis \\
3 & $\mathbf{k}\perp\hat{a}$, $\mathbf{E}\perp\hat{a}$     & field across the symmetry axis \\
\bottomrule
\end{tabular}
\caption{Orientation index of the DH21 table. $\hat{a}$ is the spheroid
symmetry axis.}
\label{tab:jori}
\end{table}

%====================================================================
\section{Derived quantities}

For a grain whose symmetry axis lies in the plane of the sky and which is
perfectly aligned, the polarization cross section is the difference the grain
presents to the two linear polarizations, halved:
\begin{equation}
Q_{\rm pol} = \tfrac{1}{2}\left[\,Q(\code{jori}{=}3) - Q(\code{jori}{=}2)\,\right],
\label{eq:qpol}
\end{equation}
formed separately from $Q_{\rm ext}$ and from $Q_{\rm abs}$, giving
$Q_{\rm pol,ext}$ and $Q_{\rm pol,abs}$. The random-orientation average of the
same table is the trace,
\begin{equation}
Q_{\rm ran} = \tfrac{1}{3}\left[\,Q_1 + Q_2 + Q_3\,\right],
\label{eq:qran}
\end{equation}
which is the modified picket fence approximation (MPFA; \S3.2 of the theory
report). Cross sections follow from
\begin{equation}
C = Q\,\pi a_{\rm eff}^{2},
\end{equation}
with $a_{\rm eff}$ converted to cm. The alignment efficiency is the
Hensley \& Draine (2023) Table 1 fit,
\begin{equation}
\falign(a) = \frac{f_{\rm max}}{1 + (a_{\rm align}/a)^{\alpha}},
\qquad
a_{\rm align} = 0.0749\um,\;\;
\alpha = 1.80,\;\;
f_{\rm max} = 1.00 .
\label{eq:falign}
\end{equation}

%====================================================================
\section{Module map}

Table~\ref{tab:modules} lists what each file is responsible for. Nothing else
in the tree touches polarization.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}p{0.30\textwidth}p{0.63\textwidth}@{}}
\toprule
File / entity & Responsibility \\
\midrule
\code{sed/src/q\_table\_jori.f90} &
The reader. Decompresses, parses the three quantity blocks, validates
(monotone axes, finite values, no trailing data), forms
Eqs.~(\ref{eq:qpol})--(\ref{eq:qran}), and supplies \code{falign\_hd23(a)},
Eq.~(\ref{eq:falign}). Publishes \code{qpol\_ext}, \code{qpol\_abs} and
\code{qran\_*}; the orientation-resolved \code{qext\_j}/\code{qabs\_j}/%
\code{qsca\_j} are \code{private} and are deallocated once the combinations
have been formed. \\[2pt]
\code{sed\_astrodust.f90}: \code{build\_Cpol} &
Interpolates $Q_{\rm pol}$ from the 169-node table axis onto the
size-distribution grid in $\log a$, exactly as \Cabs\ and $C_{\rm sca}$ are,
multiplies by $\pi a_{\rm eff}^2$, and fills \code{falign\_ad}. Checks that the
polarized and $Q$ wavelength grids agree node for node. \\[2pt]
\code{sed\_astrodust.f90}: \code{Cpol}, \code{Cpol\_ext}, \code{falign\_ad} &
Module-level arrays, \code{(NLAM,NA)} and \code{(NA)}. \code{Cpol} is the
\emph{absorption} one and drives polarized emission; \code{Cpol\_ext} is its
extinction twin and drives dichroic extinction. \\[2pt]
\code{dust\_model\_mod.f90}: \code{grain\_pop\_t} &
Carries \code{Cpol(:,:)} and \code{falign(:)}. Both unallocated is how the
solver recognizes an unpolarized population. Also carries the extinction-side
\code{Cpol\_ext(:,:)} and \code{gsca(:,:)}, read only by
\code{dust\_extinction}. \\[2pt]
\code{sed\_astrodust.f90}: \code{set\_pop} &
Optional \code{Cpol\_in}, \code{falign\_in}. Both must be supplied together;
supplying neither leaves the population unpolarized. The extinction-side
\code{Csca\_in}, \code{Cpol\_ext\_in} and \code{gsca\_in} are optional and
independent; a population that omits them contributes zero to those terms of
the size integral. \\[2pt]
\code{sed\_astrodust.f90}: \code{sed\_grain\_loop} &
Optional \code{Cpol\_pop}, \code{falign\_pop}, \code{Jpol}, again all three or
none (\code{do\_pol}). Accumulates the polarized channel alongside \code{Jout}
in every solver branch (\S\ref{sec:sites}). \\[2pt]
\code{sed\_astrodust.f90}: \code{dust\_emission} &
Optional \code{lamI\_pol}, last argument. Sums \code{Jpol} over the populations
that carry polarized optics and applies the same unit convention as
\code{lamI\_total}. \\[2pt]
\code{sed\_astrodust.f90}: \code{dust\_extinction} &
The extinction accessor. Integrates \Cabs, $C_{\rm sca}$, the
scattering-weighted $\langle\cos\theta\rangle$ and
$C_{\rm pol,ext}\,\falign$ over the size distribution of every population, and
returns them per H atom on \code{m\%lam}. These are the same numbers column 8
of the table carries, without the file (\S\ref{sec:whyaccessor}). \\[2pt]
\code{sed/src/calc\_polext.f90} &
Standalone check: computes the polarized extinction directly from
\code{qpol\_ext} and the release size distribution, and compares against the
published \code{polarized\_extinction.dat}. Independent of \code{sed\_init}. \\[2pt]
\code{sed/src/calc\_kext\_astrodust.f90} &
Writes column 8, \code{C\_polext/H}, of \code{data/kext\_astrodust\_MW.dat} as
$\sum_a {\rm d}n_{\rm Ad}(a)\,C_{\rm pol,ext}(\lambda,a)\,\falign(a)$. \\
\bottomrule
\end{tabular}
\caption{Responsibility of each file touched by the polarization update.}
\label{tab:modules}
\end{table}

%====================================================================
\section{Implementation decisions}

This is the part worth reading again in six months.

\subsection{The accumulation sites, and the two that hide}\label{sec:sites}

\code{sed\_grain\_loop} branches on \code{stoch\_method} into four cases, and
the polarized channel has to be accumulated wherever the total is. There are
eight such statements (Table~\ref{tab:sites}). Missing any one of them produces
a polarized spectrum that is silently too low rather than an error.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
branch & context & accumulates into & pattern \\
\midrule
\code{'draine'}    & equilibrium grain            & \code{Jpol\_local} & thread-local \\
\code{'draine'}    & stochastic, sum over $P(T)$  & \code{Jpol\_local} & thread-local \\
\code{'heuristic'} & equilibrium grain            & \code{Jpol}        & \textbf{direct} \\
\code{'heuristic'} & stochastic, sum over $P(T)$  & \code{Jpol}        & \textbf{direct} \\
\code{'qm'}        & QM solve, rescaled           & \code{Jpol\_local} & thread-local \\
\code{'qm'}        & GD fallback after QM failure & \code{Jpol\_local} & thread-local \\
\code{'qm'}        & equilibrium grain            & \code{Jpol\_local} & thread-local \\
\code{'equil'}     & every grain                  & \code{Jpol}        & direct \\
\bottomrule
\end{tabular}
\caption{Where the polarized emission is accumulated. The two boldface rows
are the ones a search for the \code{Jout\_local} pattern does not find.}
\label{tab:sites}
\end{table}

\paragraph{The lesson.} The \code{'draine'} and \code{'qm'} branches are
OpenMP-parallel and accumulate into a thread-local \code{Jout\_local} that is
reduced at the end. The \code{'heuristic'} branch is serial and accumulates
into \code{Jout} directly. A survey of the accumulation sites conducted by
searching for \code{Jout\_local} therefore returns only the parallel branches,
and an initial pass over this code identified four sites on exactly that basis.
Since \code{'heuristic'} is the \emph{default} solver, that omission would have
made the polarized emission come back identically zero in the default
configuration while every other solver looked correct. Any future change that
adds a term to the emission sum must be checked against
Table~\ref{tab:sites} branch by branch, not by pattern search.

\subsection{Why the \code{'qm'} rescaling is exact}

\code{qm\_solve\_grain} does not return a temperature distribution; it returns
the emitted spectrum with \Cabs\ already folded in. The polarized counterpart
is therefore obtained by rescaling:
\begin{lstlisting}
if (Cabs_pop(ii, ir) > 0.0_wp) then
   Jpol_local(ii) = Jpol_local(ii) + &
      wpol * emission_qm(ii) * (Cpol_pop(ii, ir) / Cabs_pop(ii, ir)) / &
      (4.0_wp * PI * lam(ii) * 1.0e-3_wp)
end if
\end{lstlisting}
This is not an approximation. The temperature distribution $P(T)$ is set by the
\emph{heating}, which is governed by \Cabs\ and is unchanged by the question of
which polarization is being read out. Only the emission readout changes, so
\begin{equation}
C_{\rm pol}\!\int\! P(T)\,B_\lambda(T)\,{\rm d}T
=\frac{C_{\rm pol}}{C_{\rm abs}}\;
 C_{\rm abs}\!\int\! P(T)\,B_\lambda(T)\,{\rm d}T
=\frac{C_{\rm pol}}{C_{\rm abs}}\;\code{emission\_qm}
\end{equation}
holds identically, wavelength by wavelength. Wavelengths with zero \Cabs\ emit
nothing and are skipped rather than divided by zero.

\subsection{The $P(T)$ solver was not touched}

No change was made to \code{calc\_P}, \code{narrow\_iterative}, or
\code{qm\_solve\_grain}. This is deliberate and it is what makes stochastic
heating automatically correct for polarization: at every site the same
temperature weight multiplies both accumulators, one with
${\rm d}n\,\Cabs$ and the other with ${\rm d}n\,\falign\,\Cpol$. The polarized
spectrum is thus the same average over the same $P(T)$, and no separate
stochastic treatment of the polarized channel is needed (see \S5.7 of the
theory report for why this is the physically right structure).

\subsection{Mixed orientation-average conventions}\label{sec:mixed}

This is the one open physical wart, and it is documented rather than hidden.

\Cabs\ in SEDust comes from our own T-matrix run, whose random-orientation
average is computed exactly. \Cpol\ comes from the release table, whose
available average is the $1/3$ trace, Eq.~(\ref{eq:qran}), i.e.\ the MPFA. The
two are different approximations to the same quantity, and the code mixes them:
\code{build\_Cpol} adds the polarized channel and deliberately does \emph{not}
recompute \Cabs.

Table~\ref{tab:mixed} gives the measured disagreement,
$|Q_{\rm abs}^{\rm trace}/Q_{\rm abs}^{\rm exact}-1|$, on the shared
$(\lambda, a_{\rm eff})$ grid.

\begin{table}[h]
\centering
\begin{tabular}{@{}lrr@{}}
\toprule
band & median & maximum \\
\midrule
far infrared ($\lambda > 30\um$)     & 0.022\% & 3.46\% \\
\quad restricted to $a \le 0.5\um$   &         & 1.98\% \\
mid infrared                         & 0.033\% &        \\
near infrared                        & 0.038\% &        \\
optical                              & 0.080\% &        \\
ultraviolet                          & 0.21\%  & 9.53\% \\
\bottomrule
\end{tabular}
\caption{Disagreement between the trace average and the exact
orientation average of $Q_{\rm abs}$.}
\label{tab:mixed}
\end{table}

\paragraph{Reading the far-infrared maximum.} The 3.46\% worst case occurs at
$a = 4.2\um$, a size carrying ${\rm d}n/n_{\rm H} \sim 2.8\times10^{-36}$, so
it is numerically irrelevant. Restricting to $a \le 0.5\um$, which covers
99.999\% of the geometric cross section, the worst case drops to 1.98\%.
At $a \ll \lambda$ both averages approach the same Rayleigh limit, which is why
the far infrared is well behaved.

\paragraph{Verdict: deliberately deferred in this release.} The mixture is
left in place for v1.20, as a decision rather than an oversight. The reasoning
is recorded here so that it is not rediscovered later. The mixture is harmless
for polarized \emph{emission},
which is a far-infrared phenomenon. It is \textbf{not} defensible for
polarized \emph{extinction} in the ultraviolet or the optical, and column 8
spans the whole wavelength grid. The \code{Cpol\_ext} output of
\code{dust\_extinction} is the same quantity by the same route and carries the
same caveat unchanged. Anyone using either shortward of the near
infrared must resolve this first. The clean resolution is to recompute \Cabs\
on the MPFA convention so that both channels share one average. That changes
the existing SEDs, so it breaks their byte-identity with the current reference
outputs and has to be done as a separate, deliberately validated step rather
than folded into this one. The size of the change is not an argument for
leaving it: the correctness argument is that two channels of one calculation
should not use two different orientation averages.

What was decided for this release is only \emph{when}, not \emph{whether}.
The planned use of these optics --- redoing Seon (2018) in the $V$ band, with
one or two further optical bands possible --- lies entirely in the range where
the disagreement is $0.080\%$ or below, while unifying the convention would
change \Cabs\ and break the byte-identical regression baseline that the rest
of v1.20 was validated against. Spending that baseline buys nothing inside the
present scope. It has to be spent before any ultraviolet polarized extinction
is quoted, and that is the trigger for doing the work.

\subsection{\falign\ is the fit, not the release column}

\code{data/release/size\_distribution.dat} also ships an alignment column. It
agrees with Eq.~(\ref{eq:falign}) to a median ratio of 1.0000 but differs by up
to 0.32\% at the small end. The analytic form is the canonical one in SEDust
because it is the published fit, because it is defined on any radius grid
rather than only the tabulated one, and because it keeps the alignment model in
a single place. Do not mix the two sources: a quantity computed half from one
and half from the other is neither.

\subsection{Names and surface area}

\paragraph{\code{Cpol} was not renamed to \code{Cpol\_abs}.} The name appears at
roughly 40 sites and is part of the public surface
(\code{grain\_pop\_t\%Cpol}, the optional \code{set\_pop(..., Cpol\_in=...)}).
A rename would touch every one of them for no functional gain. The declaration
carries an explicit note instead: \code{Cpol} is the absorption one,
\code{Cpol\_ext} its extinction twin.

\paragraph{The polarized extinction was exported as a table column first.} A
dedicated \code{dust\_cpol\_ext} accessor was considered at the time and
rejected: \code{dust\_lib} was then an emission-facing API with no extinction
surface at all, so such an accessor would have been the module's only
extinction entry point, a worse inconsistency than the inconvenience it
removed. The condition attached to that decision was that \code{Cpol\_ext}
should join a full extinction API if one ever appeared. It has;
see \S\ref{sec:whyaccessor}.

\subsection{Why the accessor exists now}\label{sec:whyaccessor}

The table-only export was right for a library that emitted and did not
extinguish. Rebuilding a radiative transfer code on top of SEDust changed the
premise: the host would have obtained its emission from a call and its opacity
from a parsed text file, and that asymmetry is the whole problem. The file is
also written on whatever wavelength grid the driver used, so a host on any
other grid has to interpolate optics that the model could have evaluated
exactly.

\code{dust\_extinction} closes it. The model object already held everything
required, so the accessor adds no physics and no new data source; it performs
the same size integral the driver performs, over the populations already
built. Two fields were added to \code{grain\_pop\_t} to make that possible:
\code{Cpol\_ext}, previously a module array only, and \code{gsca}, the
scattering asymmetry, which the emission path never needed and which
\code{sed\_init} now interpolates onto the size grid beside \Cabs\ and
$C_{\rm sca}$, from the same $Q$ table and by the same routine the driver used.
Agreement with the table is at the precision the file is written with (a median
relative deviation of $\sim7\times10^{-9}$ in each cross section), which is the
expected result for one integral evaluated twice.

The table remains, unchanged byte for byte, and remains the right route for a
tool that wants the optics without linking the library.

\subsection{Compression handling, and a scratch file that leaked}

The table ships compressed and Fortran cannot read a deflate stream. Adding a
zlib binding was rejected as too heavy for one file. The reader instead shells
out once,
\begin{lstlisting}
call execute_command_line('gzip -dc "'//trim(gz_file)//'" > "'// &
                          trim(out_file)//'"', exitstat=estat, cmdstat=cstat)
\end{lstlisting}
writes \code{q\_jori\_scratch.dat} in the working directory, reads it, and
deletes it. A fresh clone works with no manual setup step, and neither a 17\,MB
untracked sibling in \code{data/dielectric} nor a new build dependency is
created. A path not ending in \code{.gz} is opened directly.

\paragraph{The bug this had.} Shell redirection creates the target file
\emph{before} \code{gzip} runs, so when decompression failed the early return
left a zero-length \code{q\_jori\_scratch.dat} behind in the user's working
directory. The failure path now calls \code{discard\_scratch\_copy} before
bailing out, like every other error exit. The general shape of the bug, an error
path that skips the removal the success path performs, is worth watching for in any
future edit to this reader: it has eight separate early returns.

%====================================================================
\section{Verification}

\subsection{Polarized extinction against the release}

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
\toprule
Check & Result \\
\midrule
Column 8 vs.\ released \code{polarized\_extinction.dat} &
median 0.0296\%, max 0.9422\%, 992 points \\
Column 8 vs.\ \code{calc\_polext.x} &
agree to $7\times10^{-8}$ relative \\
Columns 1--7 of \code{kext\_astrodust\_MW.dat} &
byte-identical to the pre-change file \\
\code{astrodust\_irem\_ours\_S1/S2/PAH.dat} &
byte-identical \\
\code{dl07\_sed\_ours\_mw31\_60.dat} &
byte-identical \\
Build & zero warnings \\
\bottomrule
\end{tabular}
\caption{Verification of the polarized extinction path and regression of
everything else.}
\label{tab:verif}
\end{table}

The comparison against the release is restricted to $\lambda \ge 0.11\um$
because the reference file turns negative below that. The 0.9422\% maximum
deviation is not a systematic offset: it sits at $\lambda = 0.112\um$, right at
the sign reversal of the polarized extinction, where the quantity passes
through zero and a relative deviation is not a meaningful measure. Away from
that point the agreement is at the level of the median.

The $7\times10^{-8}$ figure is the precision the ASCII output format carries,
so column 8 and \code{calc\_polext.x} agree as closely as the files allow.
They are two independent routes to the same quantity: \code{calc\_polext.x}
works straight from \code{qpol\_ext} and the release size distribution, while
column 8 goes through \code{sed\_init}, the $\log a$ interpolation onto the SED
size grid, and \code{build\_Cpol}.

\subsection{Polarized emission}

Intrinsic polarization fraction \code{lamI\_pol/lamI\_total} under a Mathis
ISRF at $U = 1.585$:

\begin{table}[h]
\centering
\begin{tabular}{@{}lr@{}}
\toprule
$\lambda$ & fraction \\
\midrule
12\um                       & 0.00\% \\
100\um                      & 13.67\% \\
154\um                      & 17.15\% \\
350\um                      & 18.79\% \\
850\um                      & 19.15\% \\
median over 300--3000\um    & 19.18\% \\
\bottomrule
\end{tabular}
\caption{Intrinsic polarization fraction of the emitted spectrum, before any
geometric factor.}
\label{tab:polfrac}
\end{table}

The mid-infrared zero is correct rather than a failure: there the emission is
dominated by stochastically heated PAHs, which HD23 take to be unaligned.

\paragraph{19.15\% versus Planck's 22\% is a property of the model, not a bug.}
Planck measures $p_{\rm max} = 22.0\,(+3.5,-1.4)\%$ at 353\,GHz. The 19.15\%
here is the \emph{maximum} attainable value, obtained before any geometric
factor, and every real line of sight reduces it, so the model as implemented
cannot reach the observed maximum. This traces to the dust model itself: the
axial ratio $b/a = 1.4$ is the minimum non-sphericity DH21b found the starlight
polarization efficiency to require, and it was not tuned to maximize the
submillimeter polarization fraction. Do not close the gap by adjusting
\falign\ or the axial ratio without recognizing that as a change to the dust
model, requiring its own revalidation against the extinction and emission
constraints HD23 fit.

%====================================================================
\section{First-principles orientation-resolved optics}\label{sec:firstprinciples}

The table of \S\ref{sec:source} is Draine \& Hensley's precomputed one. SEDust
can now regenerate the same orientation-resolved cross sections from the
astrodust dielectric function with its own T-matrix engine, so the polarized
optics no longer have to be taken on trust from the release file. The
random-orientation optics were already SEDust's own (the \code{q\_astrodust}
table read by \code{q\_table}); this closes the remaining gap, the
orientation-resolved part read by \code{q\_table\_jori}.

Recall the three orientations of the incident wave relative to the spheroid
symmetry axis $\hat a$: $\mathrm{jori}=1$ is $\mathbf{k}\parallel\hat a$;
$\mathrm{jori}=2$ is $\mathbf{k}\perp\hat a$ with $\mathbf{E}\parallel\hat a$;
$\mathrm{jori}=3$ is $\mathbf{k}\perp\hat a$ with $\mathbf{E}\perp\hat a$. The
polarized and mean combinations are
$Q_{\rm pol}=\tfrac12(Q_3-Q_2)$ and $Q_{\rm ran}=\tfrac13(Q_1+Q_2+Q_3)$.

The optics are reached two ways. The \emph{table} is loaded once and
interpolated: this is the production path, fast and safe to call from many
threads. The \emph{direct} routine \code{oriented\_cross\_sections} computes one
node from first principles; it is the shared core the table generator loops
over, and it also serves an arbitrary $(a,\lambda)$ or a shape or porosity for
which no table exists. It must not be called from inside a loop over
radiative-transfer cells: each call is a full T-matrix solve plus an angular
integral, and the Mishchenko routines carry their state in \code{COMMON} blocks
and are not thread-safe.

\subsection{Rayleigh regime, $x<0.1$}

In the electrostatic limit the oblate spheroid has a diagonal polarizability
$\boldsymbol\alpha=\mathrm{diag}(\alpha_a,\alpha_b,\alpha_b)$, with $\alpha_a$
for $\mathbf{E}$ along the symmetry axis and $\alpha_b$ for $\mathbf{E}$
transverse, set by the depolarization factors $L_a,L_b$ of the spheroid through
$\alpha_i\propto(\varepsilon-1)/[\,1+(\varepsilon-1)L_i\,]$. Absorption and
scattering follow as $Q_{\rm abs}\propto\mathrm{Im}\,\alpha$ and
$Q_{\rm sca}\propto|\alpha|^2$. The three orientations sample exactly these two
polarizabilities: $\mathrm{jori}=1$ and $3$ present the transverse $\alpha_b$
(the field is transverse to $\hat a$ in both), $\mathrm{jori}=2$ presents
$\alpha_a$. So the orientation-resolved cross sections are the two polarization
components the random average $(\alpha_a+2\alpha_b)/3$ is already built from;
extracting them before the average is analytic and exact, and $Q(\mathrm{jori}=1)$
equals $Q(\mathrm{jori}=3)$ identically. \code{rayleigh\_limit} returns them
through optional arguments, leaving its averaged outputs untouched.

\subsection{T-matrix regime, $0.1\le x\le 50$}

The T-matrix $\mathbf{T}$ is a property of the particle alone --- size, shape,
refractive index --- independent of orientation and of the incidence and
scattering directions. SEDust solves it once for each wavelength and radius with
Mishchenko's \code{tmd\_one\_scatmat}. Two orientation-resolved quantities are
then read off the same $\mathbf{T}$.

\paragraph{Extinction, from the optical theorem.} Mishchenko's \code{AMPL}
evaluates the fixed-orientation amplitude matrix
$\mathbf{S}(\mathbf{k}\!\to\!\mathbf{k}')$ from the stored $\mathbf{T}$ for a
particle placed at chosen Euler angles. Taking the scattering direction equal to
the incidence direction (forward) and applying the optical theorem,
$C_{\rm ext}=(4\pi/k)\,\mathrm{Im}\,S_{\rm fwd}$ for the incident polarization,
gives $C_{\rm ext}$ for each orientation: the symmetry axis is placed along
$\mathbf{k}$ ($\mathrm{jori}=1$) or across it ($\mathrm{jori}=2,3$) and the
co-polar forward amplitude is read ($S_{\theta\theta}$ for $\mathbf{E}$ along
$\hat\theta$, $S_{\phi\phi}$ for $\hat\phi$).

\paragraph{Scattering, from the phase-matrix integral.} The same \code{AMPL},
swept over all scattering directions at fixed incidence, gives the differential
scattering cross section $|\mathbf{S}|^2$; the scattering cross section is its
integral over the sphere,
$C_{\rm sca}=\int(|S_{\theta\leftarrow p}|^2+|S_{\phi\leftarrow p}|^2)\,
d\Omega$ for incident polarization $p$, and absorption follows by conservation,
$C_{\rm abs}=C_{\rm ext}-C_{\rm sca}$. The fixed-orientation scattered intensity
is a trigonometric polynomial of degree ${\sim}2N_{\max}$ in $\cos\theta$ and in
$\phi$, so Gauss--Legendre nodes in $\cos\theta$ ($N_{\max}+2$ of them) and a
uniform azimuth grid ($2N_{\max}+2$ points) integrate it to machine precision.

\paragraph{An engine subtlety.} Mishchenko's random-orientation expansion
\code{GSP} reuses the \code{/TMAT/} storage as scratch through an
\code{EQUIVALENCE} and so destroys the T-matrix it reads. The amplitude
evaluation needs that same T-matrix afterward, so \code{tmd\_one\_scatmat} keeps
an intact copy in a second common block and restores \code{/TMAT/} after
\code{GSP}. Without this the oriented cross sections came out as garbage
(negative $C_{\rm ext}$). This corrects a premise stated in the planning note,
that the converged T-matrix simply remains available after the solve.

\subsection{Geometric-optics regime, $x>50$}

Here the T-matrix does not converge and an asymptotic model is used. Extinction
is the extinction paradox, $C_{\rm ext}\to 2A_{\rm proj}$, twice the geometric
shadow. For an oblate spheroid of equatorial semi-axis $a_s$ and symmetry
semi-axis $c$ ($a_s/c=b/a$), the shadow is $\pi a_s^2$ when $\mathbf{k}$ is along
the symmetry axis ($\mathrm{jori}=1$) and $\pi a_s c$ when it is across
($\mathrm{jori}=2,3$), giving $Q_{\rm ext}(1)=2(b/a)^{2/3}$ and
$Q_{\rm ext}(2)=Q_{\rm ext}(3)=2(b/a)^{-1/3}$. The shadow depends on the
direction of $\mathbf{k}$ relative to the axis but not on the incident
polarization, so extinction carries no polarization at this order.

The polarization in this regime is carried by absorption. In the opaque limit,
where a refracted ray is fully absorbed --- valid when
$\mathrm{Im}(m)\,x\gg1$ --- the absorptivity of an illuminated surface element is
one minus its Fresnel power reflectance, and that reflectance is
polarization-dependent ($R_s\neq R_p$ away from normal incidence). Integrating
over the illuminated half of the spheroid,
\[
  C_{{\rm abs},p}=\int_{\rm illum}
    \bigl[\,1-\bigl(|\hat e\!\cdot\!\hat s|^2 R_s(\theta_i)
                   +|\hat e\!\cdot\!\hat p|^2 R_p(\theta_i)\bigr)\bigr]
    \,dA_{\rm proj},
\]
where $\theta_i$ is the local angle of incidence and $\hat s,\hat p$ the local
senkrecht and parallel directions. The two grain-frame polarizations
($\hat e=\hat a$ versus $\hat e\perp\hat a$) project differently onto the local
$\hat s,\hat p$ basis as they sweep the curved surface, and that is the whole
source of the dichroism. A purely geometric ray treatment, with no Fresnel
coefficients, would give no polarization at all; the surface reflectance is what
makes the regime dichroic, and it is needed to match the release, which carries
a real absorption dichroism there --- about $3\%$ of the mean, and
\emph{growing} with size parameter, the wave-optics signature a ray limit cannot
reproduce. This is a deliberate approximation, documented at the code site with
its opaque-limit domain of validity; it would matter for a size distribution
with large-grain weight, which the astrodust distribution does not have.

\subsection{Dispatch, generation, and drop-in}

\code{oriented\_cross\_sections} selects the regime by $x$ and returns the
oriented $(Q_{\rm ext},Q_{\rm abs},Q_{\rm sca})$ for one node, with the same
branch boundaries and fallbacks as the random-orientation driver. \code{run\_q\_jori.x}
loops it over the DH21 grid and writes the table in the exact stream the release
reader expects (\S\ref{sec:source}), so the result is a drop-in:
\code{sed\_init} and \code{build\_astrodust} already take an optional
\code{qpol\_path} that defaults to the release file, and pointing it at the
regenerated table switches the polarized optics to SEDust's own with no code
change. A full sweep is dominated by the T-matrix nodes (of order 16~hours on
one core), so the driver has a wavelength-window \code{range} mode for
process-level parallelism and a \code{merge} mode that reassembles the windows
--- a window is not a contiguous slice of the stream, because the wavelength
index runs innermost.

\subsection{How far it has been checked}\label{sec:fpcheck}

Each regime was compared node by node against the release table
(Table~\ref{tab:fpcheck}). Across the Rayleigh and T-matrix regimes --- every
wavelength and radius that carries astrodust weight --- the reconstruction
matches the release to a median of a few parts in $10^4$, in both the mean and
the polarized cross sections. The geometric-optics region agrees only coarsely,
as an asymptotic model should: the mean absorption to about ten percent and the
dichroism to about three-quarters of the release value, both with the correct
sign and size-parameter trend. It joins the T-matrix continuously across $x=50$
in sign and ordering, and it carries negligible weight in the size integral.

End to end, feeding the regenerated table through \code{calc\_polext} reproduces
the released polarized extinction to a median of $0.06\%$ over
$\lambda\ge0.11\um$, against $0.03\%$ for the release table itself; the two agree
with each other to a median of $0.02\%$ at the level of the observable. So the
few-parts-in-$10^4$ reconstruction error at the level of a single cross section
survives the size integral as a few parts in $10^4$ in the polarized extinction,
computed entirely from the astrodust dielectric function with no recourse to the
release optics.

\begin{table}[h]
\centering
\small
\begin{tabular}{llll}
\toprule
Regime & $x$ & Quantity & median $|{\rm ours}/{\rm HD23}-1|$ \\
\midrule
Rayleigh  & $<0.1$      & $Q_{\rm ran}$ (abs) & $2.3\times10^{-4}$ \\
          &             & $Q_{\rm pol}$ (abs) & $3.7\times10^{-4}$ \\
T-matrix  & $0.1$--$50$ & $Q_{\rm ran}$ (ext) & $2.1\times10^{-4}$ \\
          &             & $Q_{\rm pol}$ (ext) & $8.4\times10^{-4}$ \\
          &             & $Q_{\rm ran}$ (abs) & $1.9\times10^{-4}$ \\
          &             & $Q_{\rm pol}$ (abs) & $7.6\times10^{-4}$ \\
GO        & $>50$       & $Q_{\rm ran}$ (abs) & $9.3\times10^{-2}$ \\
          &             & $Q_{\rm pol}$ (abs) & $3.3\times10^{-1}$ \\
\bottomrule
\end{tabular}
\caption{Node-by-node reconstruction against the HD23 release table. The
geometric-optics dichroism is at $\sim0.75$ of the release with matching sign
and trend; its size-parameter region carries no astrodust weight.}
\label{tab:fpcheck}
\end{table}

%====================================================================
\section{Aligned-grain scattering matrix and the extinction matrix}
\label{sec:aligned}

Section~\ref{sec:firstprinciples} regenerated the orientation-resolved
\emph{cross sections}. This section adds the angle-resolved object they left out:
the fixed-orientation Mueller matrix $Z(\theta_i;\theta_s,\phi)$ of the aligned
astrodust spheroid, and the $4\times4$ extinction matrix $K(\theta_i)$ that goes
with it. Between them they complete the material side of the Peest-formalism
contract --- SEDust returns every quantity in matrix form; the RT code (MoCafe,
whose consumption side is not yet built) does the rotations, sampling, peel-off,
and $\exp(-K\tau)$ transfer. The design is set out in
\code{docs/tmatrix\_aligned\_scattering\_plan.md}; this section records how each
piece was built and why.

\subsection{The engine}\label{sec:aligned:engine}

\code{tmatrix/driver/scattering\_matrix\_oriented.f90} returns the
fixed-orientation Mueller matrix $Z$ of the DH21 oblate spheroid ($b/a=1.4$) for
one incidence direction. Two regimes, one output convention.

\paragraph{T-matrix regime, one \code{AMPL} call.} With the T-matrix already
converged and left in \code{COMMON /TMAT/} by \code{TMD\_ONE\_SCATMAT} --- the
same restore of \S\ref{sec:firstprinciples} that made the oriented cross sections
possible --- a single call to Mishchenko's \code{AMPL} evaluates the $2\times2$
complex amplitude matrix $\mathbf{S}=[[S_{vv},S_{vh}],[S_{hv},S_{hh}]]$ for the
chosen incidence and scattering angles. The 16 phase-matrix elements then follow
from the standard bilinears in $\mathbf{S}$, copied \emph{verbatim} from
Mishchenko's \code{src/ampld.lp.f} (its \code{Z11}$\ldots$\code{Z44}, Eqs.\
(13)--(29) of Appl.\ Opt.\ \textbf{39}, 1026, 2000) so the phase matrix can be
checked line by line against the source. Copying rather than rederiving is
deliberate: the bilinears carry sign and conjugation conventions that are easy to
transcribe wrongly, and the whole point of anchor~C below is that this transcription
is exact.

\paragraph{Rayleigh regime, analytic dipole.} Below $x=0.1$ the same $Z$ is the
electric-dipole matrix, $S_{pq}=k^2\,(\hat e_p^{\rm sca}\!\cdot\!\boldsymbol\alpha
\!\cdot\!\hat e_q^{\rm inc})$ with the spheroid polarizability
$\boldsymbol\alpha=\mathrm{diag}(\alpha_b,\alpha_b,\alpha_a)$, then the same
bilinears. The polarizability comes from \code{spheroid\_dipole\_polarizability},
a routine extracted from \code{rayleigh\_limit} (\S\ref{sec:firstprinciples})
without changing an operation, so the dipole cross sections that
\S\ref{sec:firstprinciples} already validated and the dipole matrix here rest on
one implementation of $\alpha_a,\alpha_b$.

\paragraph{The Stokes basis, stated once.} Both regimes use Mishchenko's
meridional basis of each propagation direction \emph{in the grain frame}, whose
$z$ axis is the alignment axis: $\hat e_1=\hat\theta$ (call it $v$), $\hat
e_2=\hat\phi$ ($h$), with $Q=I_v-I_h$. This is exactly the basis \code{AMPL}
rotates its amplitudes into, so the analytic dipole and the T-matrix matrix share
one convention with no adaptor between them. At $\theta_i=90^\circ$ the
$v$-polarization is $\mathrm{jori}=2$ ($\mathbf{E}\parallel\hat a$) and $h$ is
$\mathrm{jori}=3$ ($\mathbf{E}\perp\hat a$), which is what ties the extinction
matrix back to the jori cross sections of \S\ref{sec:source}.

\subsection{Symmetries, verified not assumed}\label{sec:aligned:sym}

Three exact symmetries reduce the stored ranges and, more usefully, are numerical
checks on the engine (anchor~E). Each was confirmed on a grid of
$(\theta_i,\theta_s,\phi)$ before being relied on.

\begin{itemize}
\item \textbf{Azimuth mirror.} The plane through the axis and the incidence
      direction is a symmetry plane, so $\phi\to360^\circ-\phi$ leaves $Z$
      unchanged except that the two off-diagonal $2\times2$ blocks (elements
      $13,14,23,24,31,32,41,42$) flip sign. Store $\phi\in[0,180^\circ]$; residual
      $3.5\times10^{-7}$.
\item \textbf{Equatorial mapping.} The oblate spheroid is symmetric under
      $z\to-z$, so $\theta_i\to180^\circ-\theta_i$ maps onto the stored range with
      $\theta_s\to180^\circ-\theta_s$, $\phi$ \emph{unchanged}, and the same
      off-diagonal-block sign flip,
      \[
        Z(180^\circ-\theta_i;\,180^\circ-\theta_s,\,\phi)\;=\;S\odot
        Z(\theta_i;\,\theta_s,\,\phi),\qquad
        S=\begin{pmatrix}+&+&-&-\\ +&+&-&-\\ -&-&+&+\\ -&-&+&+\end{pmatrix},
      \]
      the diagonal blocks keeping their sign. Store $\theta_i\in[0,90^\circ]$;
      residual $8.8\times10^{-16}$. Physically this also encodes that the
      alignment axis is headless.
\item \textbf{Axisymmetry at $\theta_i=0$.} There $Z$ must be independent of
      $\phi$ up to a frame rotation, $Z(0;\theta_s,\phi)=Z(0;\theta_s,0)\,
      R(\phi)$, the $\phi=0$ matrix being six-element block-diagonal. This is a
      check, not a storage saving.
\end{itemize}

The library reconstructs the unstored octants from these relations, applying the
$\pm1$ sign once to the interpolated matrix rather than to eight interpolation
corners (\S\ref{sec:aligned:lib}); because each reflection contributes a constant
that is the same for every corner of a cell, it factors out of the linear
interpolation exactly.

\subsection{The generator: one code path}\label{sec:aligned:gen}

The shared physics lives in \code{tmatrix/driver/aligned\_population\_optics.f90},
and both the production driver \code{run\_scatmat\_aligned.f90} and the
verification program \code{compare\_scatmat\_aligned.f90} call it. This is on
purpose: the numbers an anchor checks are produced by the exact code the
production file is written from, so a passing anchor certifies the shipped path
and not a parallel one.

\paragraph{One solve, three products.} For each (band, size) in the T-matrix
regime a single \code{TMD\_ONE\_SCATMAT} solve serves everything: its \code{GSP}
expansion coefficients feed the random-orientation accumulation (the $F$ block
below), the \code{/TMAT/} it leaves valid feeds the \code{AMPL} sweep over the
$Z$ nodes, and its forward amplitudes give the extinction-matrix elements. No
quantity is recomputed.

\paragraph{Regime split and the $x>50$ guard.} The size-parameter regimes match
\code{run\_scatmat.f90}: $x<0.1$ analytic dipole, $0.1\le x\le50$ T-matrix. Above
$x=50$ no converged oriented matrix exists, so the bin is \emph{omitted from the
aligned products} and instead enters $F_{\rm tot}$ as its random-orientation
geometric-optics matrix, i.e.\ scattered as unaligned. Its $C_{\rm sca}$ is
tallied, and the driver stops hard if the omitted fraction of the scattering
weight exceeds $10^{-6}$. It never does for the shipped bands: the measured
omitted fraction is at most $3.1\times10^{-15}$ (at $0.36\um$) and exactly zero at
$0.64$ and $0.79\um$. The omission is thus provably negligible rather than
assumed so.

\paragraph{An OpenMP lesson: \code{-frecursive} for the amplitude routine.} The
only parallel region is the loop over $Z$ nodes inside one size. \code{AMPL} and
\code{VIGAMPL} only \emph{read} \code{COMMON /TMAT/} and otherwise write local
automatic arrays, so distinct nodes are independent and each grid slot is written
once --- which is why a 1-thread and an $N$-thread run are bitwise identical. The
subtlety is in the build: \code{ampl\_oriented.f} is compiled \code{-frecursive}
so its large local arrays become thread-private on the stack, while the serial
T-matrix core (\code{tmd\_one.f}, \code{lpd.f}) keeps its static locals under the
base flags. Compiling the amplitude routine with static locals would have shared
those arrays across threads and silently corrupted the matrices; the Makefile
therefore compiles that one file with the OpenMP-safe flags and the core without
them. This is the implementation lesson worth carrying forward: thread-safety of a
COMMON-based Mishchenko routine is decided by the compile flags of each file, not
by a source edit.

\subsection{Products and the $\eta$ contract}\label{sec:aligned:products}

Per band the file (\code{output/scatmat\_aligned\_astrodust\_P0.20\_Fe0.00\_1.400.dat})
carries three blocks, all per H.

\begin{itemize}
\item \textbf{$K$ block}, on the $\theta_i$ grid: $C_{\rm ext,al}$, the dichroic
      $C_{\rm pol,al}$, and the birefringent $C_{\rm bir,al}$ from the forward
      amplitudes, plus $C_{\rm sca,al}$ as the grid closure $\int Z_{11}\,
      d\Omega$. Because these come from the forward amplitude at each $\theta_i$,
      the extinction matrix is \emph{exact at every propagation angle}. This
      replaces the $\sin^2\psi$ interpolation between the $\psi=0$ and $90^\circ$
      values that the three-orientation jori table forces, and reduces to that law
      analytically in the Rayleigh limit (anchor~G).
\item \textbf{$F$ block}: $F_{\rm tot}$ (every grain, identical to
      \code{run\_scatmat.f90}) and $F_{\rm ref}$ ($f_{\rm align}$-weighted), the
      six-element random-orientation matrices, $\alpha_1$-normalized so
      $\tfrac12\int F_{11}\,d\cos\Theta=1$; the header gives $C_{\rm sca,tot}$,
      $C_{\rm sca,ref}$, $C_{\rm ext,tot}$, $C_{\rm ext,ref}$ to restore absolute
      units.
\item \textbf{$Z$ block}: $Z_{\rm al}=\int da\,n(a)\,f(a)\,Z(a)$, 16 elements,
      $\um^2\,\mathrm{sr}^{-1}$ per H, on
      $\theta_i=0(5)90^\circ$ (19) $\times$ $\theta_s=0(1)180^\circ$ (181)
      $\times$ $\phi=0(5)180^\circ$ (37).
\end{itemize}

\paragraph{The $\eta$ contract.} The file header documents, and the library
enforces, the exact rule for varying the alignment degree cell to cell by a single
scalar $\eta$ (the local alignment scale, $f_{\rm cell}=\eta f_{\rm ref}$). The
aligned matrix scales linearly, $Z_{\rm al,cell}=\eta Z_{\rm al}$; the extinction
matrix likewise, $K_{\rm al,cell}=\eta K_{\rm al}$; the unaligned remainder
scattering matrix is $F_{\rm unal}=F_{\rm tot}-\eta F_{\rm ref}$ in absolute units
(after restoring $C_{\rm sca,tot}$ and $C_{\rm sca,ref}$); and the unaligned
population adds an isotropic $C_{\rm ext,tot}-\eta C_{\rm ext,ref}$ to the
extinction (its polarized and birefringent parts vanish). All of this is exact by
the linearity of the size integral in $f_{\rm align}$, so three stored integrals
give the scattering optics of any $\eta$ with no re-integration.

\subsection{The library: two strictly separated layers}\label{sec:aligned:lib}

\code{sed/src/scatmat\_aligned.f90} (module \code{scatmat\_aligned\_mod}) serves
the table to a polarized RT host with the same lifecycle as the rest of the
library: initialize once, query along the path.

\paragraph{Initialization (serial, once).} \code{load\_scatmat\_aligned(path,
status)} parses the table into module storage (about $81.5$\,MB resident for the
production grid) and \code{free\_scatmat\_aligned} releases it. It is wired into
\code{sed\_init} / \code{build\_astrodust} through an optional \code{scatmat\_path}
argument --- exactly the \code{qpol\_path} pattern of \S\ref{sec:firstprinciples},
except that a \code{scatmat\_path} that fails to load is an error (\code{status =
3}), because a host that asked for the table depends on it, whereas a missing
polarized-optics table only disables polarized emission. This layer is not
thread-safe and is called from serial code.

\paragraph{Path queries (pure reads, OpenMP-safe).} Five routines write only their
own outputs and read the loaded arrays, so they are safe to call concurrently from
photon threads:
\begin{itemize}
\item \code{scatmat\_band(lambda\_um, iband, exact)} --- nearest stored band, so
      the hot path takes an integer and no wavelength search runs per event;
\item \code{extinction\_matrix\_aligned(iband, theta\_i, eta, kmat)} --- the
      $4\times4$ matrix with diagonal $\eta\,C_{\rm ext,al}$,
      $K(1,2)=K(2,1)=\eta\,C_{\rm pol,al}$, $K(3,4)=\eta\,C_{\rm bir,al}$,
      $K(4,3)=-\eta\,C_{\rm bir,al}$, $\theta_i$ folded into $[0,90^\circ]$ because
      $K$ is even under the equatorial reflection;
\item \code{mueller\_matrix\_aligned(iband, theta\_i, theta\_s, phi, z)} ---
      trilinear interpolation plus the symmetry reconstruction of
      \S\ref{sec:aligned:sym}, at the reference alignment ($\eta=1$; the caller
      scales);
\item \code{mueller\_matrix\_random(iband, big\_theta, f\_tot, f\_ref)} --- the two
      six-element random-orientation matrices;
\item \code{scattering\_cross\_sections(iband, theta\_i, eta, csca\_aligned,
      csca\_unaligned)} with $\code{csca\_unaligned}=C_{\rm sca,tot}-\eta\,
      C_{\rm sca,ref}$.
\end{itemize}
The loaded grids and arrays are also exposed \code{public, protected}, so a host
that prefers to inline its own lookups can take them once at startup and never
call back. The whole cell dependence enters through just two scalars the host
already holds: $\eta$, and $\theta_i=\arccos(\hat k\!\cdot\!\hat B)$ from one dot
product.

\paragraph{The alignment guard (status 4).} The $K$ and $Z$ arrays were integrated
over size once, under the profile recorded in the header; the sanctioned runtime
variation is $\eta$, not a change of profile. So when a loaded scattering table is
present and the host calls \code{dust\_set\_alignment} or
\code{dust\_set\_alignment\_profile} with a profile that differs from the recorded
one, the setter returns a \emph{new} status code~\code{4}. It is non-fatal: the
$f_{\rm align}$ update still completes (the emission and $C_{\rm pol,ext}$ channels
re-integrate live and honor the new profile), and the code only signals that the
loaded scattering matrices no longer correspond to the model's alignment. A
different profile requires regenerating the table
(\code{run\_scatmat\_aligned.x profile=FILE}), which takes minutes. A tabulated
profile never matches the recorded analytic one and so always raises the flag.

\subsection{Birefringence through \code{dust\_extinction}}\label{sec:aligned:cbir}

\code{dust\_extinction} gained an optional \code{Cbir\_ext} output, the
$f_{\rm align}$-weighted size integral of the jori table's 4th-block
birefringence,
\[
  C_{\rm bir,ext}(\lambda)=\sum_a {\rm d}n_{\rm Ad}(a)\,
     C_{\rm bir,ext}(\lambda,a)\,f_{\rm align}(a),
\]
mirroring \code{Cpol\_ext} exactly. It is zero when the loaded table carries no
4th block (as the release table does not), so the default path is unchanged, and
it is the same birefringence the aligned $K$ block carries at $\theta_i=90^\circ$,
reached here without loading the aligned scattering table. This closes the gap
noted in the previous release, that \code{dust\_extinction} did not yet return the
birefringence.

\subsection{Verification}\label{sec:aligned:verif}

Every anchor of the plan passes (Table~\ref{tab:alignverif}). The engine anchors
(A--G) run at single sizes through \code{compare\_scatmat\_aligned.x}; the library
checks run on the production table.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llp{0.30\textwidth}l@{}}
\toprule
& Check & What it certifies & Result \\
\midrule
\multicolumn{4}{@{}l}{\emph{Engine (single size)}}\\
A & closure $\int Z_{11}\,d\Omega$ vs $C_{\rm sca}(\mathrm{jori})$ & phase-matrix normalization & $2.7\times10^{-7}$ \\
B & optical theorem vs $C_{\rm ext}(\mathrm{jori})$ & forward-amplitude geometry & $2\times10^{-14}$ \\
C & orientation average vs GSP $F(\Theta)$ & the bilinears, incl.\ $F_{12},F_{34}$ signs & $1.4\times10^{-6}$ \\
D & dipole vs closed form & Rayleigh implementation & $2\times10^{-16}$ \\
  & dipole vs T-matrix at $x\!\sim\!0.1$ & regime continuity, $|\delta Z|/Z_{11}$ & $0.5$--$1.1\%$ \\
E & $\phi$ mirror & azimuth symmetry & $3.5\times10^{-7}$ \\
  & equatorial mapping & $z\to-z$ symmetry & $8.8\times10^{-16}$ \\
G & Rayleigh $K(\theta_i)$ vs $\sin^2$ law & extinction-matrix angle law (exact) & $3.3\times10^{-16}$ \\
  & T-matrix $K(\theta_i)$ vs $\sin^2$ law & at $x=0.137$ & $6.3\times10^{-4}$ \\
\midrule
\multicolumn{4}{@{}l}{\emph{Generator (size-integrated)}}\\
F & $K$ vs independent jori integrals & $C_{\rm ext}$ & $1.6\times10^{-6}$ \\
  &                                   & $C_{\rm pol}$ & $2.0\times10^{-5}$ \\
  &                                   & $C_{\rm bir}$ & $1.3\times10^{-7}$ \\
\midrule
\multicolumn{4}{@{}l}{\emph{Library (production table)}}\\
  & reader round-trip & parse vs re-emit & bitwise 0 \\
  & symmetry reconstruction & unstored octants & bitwise 0 \\
  & $\eta$ algebra & linear scaling & residual 0 \\
  & $K(90^\circ)$ vs jori vs \code{dust\_extinction} & $C_{\rm pol}$ / $C_{\rm bir}$ & $1.6\times10^{-5}$ / $8.0\times10^{-5}$ \\
  & \code{rt\_example} closures & end-to-end consumption & $0.997$--$1.001$ \\
\bottomrule
\end{tabular}
\caption{Verification of the aligned scattering matrix, its extinction matrix, and
the library API. Engine and generator anchors are those of the plan's \S5; the
library checks run on the shipped production table.}
\label{tab:alignverif}
\end{table}

\paragraph{Cost of the production run.} The five UBVRI bands (0.36, 0.44, 0.55,
0.64, 0.79\um) take about 8 minutes at 32 threads (about 26 minutes at 8); the
file is 138\,MB, 38\,MB gzipped. The existing regressions are unchanged:
\code{calc\_polext} reproduces the release to a median of $0.0296\%$ as before,
and every \code{sed}, \code{mc} and \code{tmatrix} target builds clean.

%====================================================================
\section{Limits and extension paths}

These are limits of the published optics, not of the code.

\paragraph{Birefringence is now computed; the release table has none.} The
release table stores $Q_{\rm ext}$, $Q_{\rm abs}$ and $Q_{\rm sca}$ only, no phase
retardation, so from it $C_{\rm circ}=0$. The first-principles route of
\S\ref{sec:firstprinciples} already evaluates the fixed-orientation amplitude
matrix (Mishchenko's \code{AMPL}, in \code{tmatrix/src/ampl\_oriented.f}) and used
only its imaginary forward part, for extinction. Keeping the real part as well
costs nothing --- it comes from the same amplitude call --- and gives the
birefringence, the phase retardation that couples $U$ and $V$. It is stored as an
optional 4th block of the regenerated table, $Q_{\rm re}(\mathrm{jori})=
(4\pi/k)\,\mathrm{Re}[S_{\rm fwd}]/(\pi a^2)$, the real twin of $Q_{\rm ext}$, from
which $C_{\rm bir}=\tfrac12(Q_{{\rm re},3}-Q_{{\rm re},2})\,\pi a^2$. The reader
exposes it as \code{qbir\_ext} with a \code{has\_bir} flag that stays \code{.false.}
for an older 3-block table, so nothing about the default path changes. No astrodust
reference for the birefringence exists, so it is certified internally by
Kramers-Kronig: $\mathrm{Re}$ and $\mathrm{Im}$ of the forward-amplitude difference
are a Hilbert pair, and the directly computed birefringence agrees with the Hilbert
transform of the already-validated dichroism to a median $\sim0.1\%$ over the
interior of the grid (the transform itself self-tests to $2\times10^{-5}$ on a
Lorentz oscillator; the residual is the finite-range truncation). What remains is to
return it through \code{dust\_extinction} and to consume it in the $U\leftrightarrow V$
transfer term, both on the host side.

\paragraph{The aligned scattering matrix is now computed.} The release gives the
integrated $Q_{\rm sca}$ with no angular information, so scattering by an aligned
spheroid could not be modeled from it. It is now modeled from first principles:
the fixed-orientation Mueller matrix $Z(\theta_i;\theta_s,\phi)$ of the aligned
astrodust spheroid is built from the same fixed-orientation amplitude
(\code{ampl\_oriented.f}) used above for extinction, size-integrated over the
astrodust distribution with the alignment weight, and paired with the
$4\times4$ extinction matrix $K(\theta_i)$ from the forward amplitudes. It is
written for five optical bands by \code{run\_scatmat\_aligned.x} and served to a
polarized RT host through a two-layer library API. \S\ref{sec:aligned} is the
full account.

\paragraph{The random-orientation scattering matrix does exist, in five optical
bands.} This is a different object from the previous paragraph and should not be
confused with it. Dropping the alignment, a randomly oriented spheroid with a
plane of symmetry has six independent scattering-matrix elements rather than
sixteen, and those \emph{are} computed in the tree: \code{tmatrix/driver/run\_scatmat.f90} restores Mishchenko's \code{MATR} expansion, accumulates
the single-size matrices over the HD23 size distribution with $C_{\rm sca}\,{\rm
d}n/n_{\rm H}$ weighting, and writes
\code{tmatrix/output/scatmat\_astrodust\_P0.20\_Fe0.00\_1.400.dat}: one block per
wavelength, 181 rows at $\theta = 0, 1, \ldots, 180^\circ$, with columns
$F_{11}, F_{22}, F_{33}, F_{44}, F_{12}, F_{34}$. It answers what an unaligned
grain population scatters; it does not answer the aligned question above.

The stored wavelengths are $0.36$, $0.44$, $0.55$, $0.64$ and $0.79\um$, roughly
\textit{UBVRI}. That set was chosen, and the full sweep deliberately not run, because
the planned use is the $V$ band first with one or two further optical bands
possible, and neither the far infrared nor the ultraviolet is in scope. The
whole 1129-point grid costs about 1.7 hours and would be spent almost entirely
on wavelengths nothing is waiting for. Extending is cheap and needs no code
change: \code{./run\_scatmat.x all} does the full grid, or a list of
wavelengths does a subset, at about $5.5\,$s per wavelength.

\paragraph{Far infrared needs neither.} In that regime the albedo is
essentially zero, so the transfer requires only the extinction matrix and the
emission vector, both fully determined by what is exported. The HAWC+ 154\um\
comparison lies inside this regime, so the far-infrared milestone is reachable
with the present data and should be attempted before any scattering work.

\paragraph{Astrodust only.} DL07 and Zubko have no polarized optics.
\code{build\_dl07} releases any polarized arrays left by a previous astrodust
initialization, so \code{lamI\_pol} returns zeros rather than stale values.

\subsection{Cost of \code{build\_Cpol}}

Loading the orientation-resolved table is the most expensive part of model
construction (Table~\ref{tab:cost}).

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
\toprule
Item & Size \\
\midrule
compressed file          & 4.1\,MB \\
scratch file when expanded & 17.2\,MB \\
raw arrays \code{qext\_j}, \code{qabs\_j}, \code{qsca\_j} & 13.7\,MB, transient \\
derived \code{qpol\_*}, \code{qran\_*} & 7.6\,MB, kept \\
resident after \code{sed\_init} & 7.6\,MB \\
time                     & about 1\,s of the 3.1\,s model build \\
\bottomrule
\end{tabular}
\caption{Cost of loading the orientation-resolved table. The
orientation-resolved arrays exist only between the read and the formation
of the derived combinations.}
\label{tab:cost}
\end{table}

The orientation-resolved arrays \code{qext\_j}, \code{qabs\_j} and
\code{qsca\_j} are deallocated as soon as the combinations of
Eqs.~(\ref{eq:qpol})--(\ref{eq:qran}) have been formed. They are intermediate
quantities: nothing downstream of \code{build\_Cpol} reads them, and every
later step uses the five derived arrays only. Releasing them removes
13.7\,MB, about two-thirds of what the table would otherwise hold, and leaves
7.6\,MB resident for the rest of the run. The three arrays are \code{private}
to \code{q\_table\_jori}, so no consumer can depend on their surviving the
build.

Measured on \code{main\_astrodust.x}, the maximum resident set size falls from
51232\,kB to 40088\,kB, a reduction of about 10.9\,MB. The peak does not fall by the full 13.7\,MB because it
is a high-water mark: once the orientation-resolved arrays are gone, the
later SED solution stages set the peak instead. The steady-state occupancy
just after \code{sed\_init} does drop by the full 13.7\,MB.

%====================================================================
\section{Reproducing the numbers}

From \code{SEDust/sed/}:
\begin{lstlisting}[style=sh]
make calc_polext.x           # standalone polarized-extinction check
./calc_polext.x              # -> output/polarized_extinction_ours.dat
                             #    lambda, ours, reference, ratio

make calc_kext_astrodust.x   # full-pipeline extinction table
./calc_kext_astrodust.x      # -> ../data/kext_astrodust_MW.dat, 8 columns

make libsedust.a             # library, for the emission path
\end{lstlisting}

\code{calc\_polext.x} prints the median and maximum deviation against the
release directly, restricted to $\lambda \ge 0.11\um$. Column 8 of the
extinction table should reproduce it to the ASCII precision quoted in
Table~\ref{tab:verif}; if the two disagree by more than that, the $\log a$
interpolation in \code{build\_Cpol} or the size grid it interpolates onto is
the place to look, since that is the only step separating the two routes.

For the emission fractions of Table~\ref{tab:polfrac}, build a model with
\code{build\_astrodust}, evaluate \code{J\_Mathis} at $U = 1.585$, and call
\code{dust\_emission} with \code{lamI\_pol} requested; the example in \S5.4 of
the user manual is the complete recipe.

%====================================================================
\section{References}

\begin{itemize}
\item Draine, B. T., \& Hensley, B. S. 2021b, ApJ, 919, 65 (DH21b).
      \code{2021ApJ...919...65D}
\item Hensley, B. S., \& Draine, B. T. 2023, ApJ, 948, 55 (HD23).
      \code{2023ApJ...948...55H}
\item Peest, C., Camps, P., Stalevski, M., Baes, M., \& Siebenmorgen, R.
      2017, A\&A, 601, A92.
      \code{2017A\&A...601A..92P}
\item Peest, C., et al.\ 2023, A\&A, 673, A112.
      \code{2023A\&A...673A.112P}
\item Planck Collaboration 2020, A\&A, 641, A12.
      \code{2020A\&A...641A..12P}
\item Reissl, S., Wolf, S., \& Brauer, R. 2016, A\&A, 593, A87 (POLARIS I).
      \code{2016A\&A...593A..87R}
\item Seon, K.-I. 2018, ApJ, 862, 87.
      \code{2018ApJ...862...87S}
\end{itemize}

\end{document}
