\documentclass{my_memo}

\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\sloppy

\lstset{
  language=Fortran,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue!60!black},
  commentstyle=\color{green!40!black},
  stringstyle=\color{red!60!black},
  columns=fullflexible,
  breaklines=true,
  frame=single,
  framesep=4pt,
  showstringspaces=false,
  numbers=none
}
\lstdefinestyle{sh}{language=bash,keywordstyle=\color{black},basicstyle=\ttfamily\small}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\lamIlam}{\ensuremath{\lambda I_\lambda}}
\newcommand{\Nh}{\ensuremath{N_{\rm H}}}
\newcommand{\Cabs}{\ensuremath{C_{\rm abs}}}
\newcommand{\Teq}{\ensuremath{T_{\rm eq}}}
\newcommand{\um}{\ensuremath{\,\mu\mathrm{m}}}

\title{SEDust: User Manual}
\author{Kwang-il Seon}
\date{2026 July 20}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This manual describes the model-agnostic dust thermal-emission library under
\code{SEDust/sed/}. The library computes the infrared emission spectrum of a
dust mixture for an arbitrary local radiation field, and is intended to be
linked into a Fortran 3D radiative-transfer (RT) code. It handles several dust
models as peers --- the HD23 \emph{astrodust}+PAH model, the Draine \& Li (2007)
silicate+carbonaceous model, the Zubko, Dwek \& Arendt (2004) BARE-GR-S model,
and arbitrary file-defined models --- through one object type and one emission
call. The numerically validated stochastic-heating solver is reused unchanged;
the library is an additive layer on top of it.
\end{abstract}

\tableofcontents

%====================================================================
\section{Overview}

A \emph{dust model} is represented by a single derived type,
\code{dust\_model\_t}. You build it once (per RT run), then call
\code{dust\_emission} once per RT cell with that cell's local mean intensity:

\begin{lstlisting}
use dust_lib
type(dust_model_t)    :: m
real(wp), allocatable :: J_lam(:), lamI_total(:)

call build_astrodust(m, qtab, sizedist, 200, 2.7_wp, 5.0e3_wp)   ! once
allocate(J_lam(m%NLAM), lamI_total(m%NLAM))
do icell = 1, ncells
   ! ... assemble J_lam(:) (mean intensity on the grid m%lam) for this cell ...
   call dust_emission(m, J_lam, lamI_total)                      ! per cell
   ! ... lamI_total(:) is lambda*I_lambda / N_H for the cell ...
end do
\end{lstlisting}

\paragraph{Design.} The library wraps the existing, FIR-validated solver core
(\code{sed\_grain\_loop}, \code{calc\_Teq}, \code{calc\_P}) without modifying it.
A model is a set of grain \emph{populations} (one per species and charge state),
each carrying its own size distribution, absorption cross sections, enthalpy, and
Planck-integral tables. \code{dust\_emission} loops the populations through the
solver and sums their emission into named output \emph{channels}. The same
populations carry the extinction-side optics, so \code{dust\_extinction}
(\S\ref{sec:ext}) returns the model's opacity from that one object as well, so
emission and extinction both reach the transfer code through calls, not files.

\paragraph{One active model at a time.} The module-global grids
(\code{lam}, \code{T\_first}, $\dots$) hold the \emph{currently built} model's
working set. Calling \code{build\_*} makes that model active. This is sufficient
for an RT run that uses a single dust model; to switch models, rebuild.

%====================================================================
\section{Building the library}

From \code{SEDust/sed/}:
\begin{lstlisting}[style=sh]
make libsedust.a      # the library archive (link this into your RT code)
\end{lstlisting}

The compiler is \code{gfortran} with \code{-O3 -march=native -fopenmp}
(edit the \code{Makefile} for ifort). \code{libsedust.a} is a static archive of
all library objects; the Fortran \code{.mod} files are written alongside it in
\code{SEDust/sed/}.

%====================================================================
\section{Models and builders}

Each builder fills a \code{dust\_model\_t} and defines the model's output
channels (Table~\ref{tab:models}).

\begin{table}[h]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
builder & model & channels & notes \\
\midrule
\code{build\_astrodust} & HD23 astrodust+PAH & \code{AD}, \code{PAH} & $N_C{=}417$, D16 graphite \\
\code{build\_dl07}      & Draine \& Li 2007   & \code{SIL}, \code{CARB} & $N_C{=}470$, D03 graphite \\
\code{build\_zubko}     & Zubko (ZDA 2004) BARE-GR-S & \code{PAH}, \code{GRA}, \code{SIL} & neutral PAH only \\
\code{build\_from\_files} & file-defined & \code{CH1}, \code{CH2}, $\dots$ & generic loader \\
\bottomrule
\end{tabular}
\caption{Built-in model builders. Each sets the model's own optics/abundance
parameters internally.}
\label{tab:models}
\end{table}

\noindent Signatures:
\begin{lstlisting}
call build_astrodust(m, qtable_path, sizedist_path, NT, T_lo, T_hi [, status])
call build_dl07     (m, qtable_path, sizedist_path, sd_index, U, NT, T_lo, T_hi [, status])
call build_zubko    (m, config_path, data_dir, NT, T_lo, T_hi [, status])
call build_from_files(m, descriptor_path, data_dir, NT, T_lo, T_hi [, status])
\end{lstlisting}

\begin{itemize}
\item \code{NT}, \code{T\_lo}, \code{T\_hi}: number and range [K] of the internal
      temperature grid (typical \code{200, 2.7, 5000}).
\item \code{sd\_index} (DL07): WD01 size-distribution index --- \code{7} =
      MW $R_V{=}3.1$, $q_{\rm PAH}{=}4.58\%$; \code{29/30/31} = LMC2; \code{32} = SMC.
\item \code{U} (DL07): ISRF intensity scaling (the DL07 reference uses $U{=}1$).
\item \code{data\_dir} (Zubko / files): directory holding the data files, e.g.
      \code{data/zubko/}.
\item \code{status} (optional, all builders): error report; see
      \S\ref{sec:status}. When present, a missing or malformed input file is
      reported through it and the model is not built, so an RT host can recover.
\end{itemize}

%====================================================================
\section{Computing emission}

\subsection{Single-cell exact solve}

\begin{lstlisting}
call dust_emission(m, J_lam, lamI_total [, lamI_chan] [, status])
\end{lstlisting}
\begin{itemize}
\item \code{J\_lam(m\%NLAM)}: the local \emph{mean intensity} on the grid
      \code{m\%lam} [\um]. For a scaled Mathis ISRF use
      \code{call J\_Mathis(U, m\%lam, J\_lam)} from module \code{radfield}.
\item \code{lamI\_total(m\%NLAM)}: the summed SED, \lamIlam$/\Nh$
      [\,erg\,s$^{-1}$\,cm$^{-2}$\,sr$^{-1}$\,H$^{-1}$\,] --- the HD23 convention.
\item \code{lamI\_chan(m\%NLAM, m\%n\_channel)} (optional): the SED of each channel.
\item \code{status} (optional): error report; see \S\ref{sec:status}.
\end{itemize}

\subsection{Solver choice (\code{m\%stoch\_method})}

The field \code{m\%stoch\_method} selects how each grain is heated
(Table~\ref{tab:solver}); \code{dust\_emission} honors it.

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
\code{m\%stoch\_method} & method & note \\
\midrule
\code{'heuristic'} & GD look-ahead T-window narrowing & \textbf{default}; fastest, serial \\
\code{'draine'}    & Guhathakurta \& Draine (1989) & reference GD \\
\code{'qm'}        & energy-space transition matrix (thermal-discrete) & most detailed \\
\code{'equil'}     & single-grain equilibrium (no stochastic) & see \S\ref{sec:eqtemp} \\
\bottomrule
\end{tabular}
\caption{Stochastic-heating solver options. The first three resolve the grain
temperature \emph{distribution} (PAH/NIR features); \code{'equil'} forces
equilibrium.}
\label{tab:solver}
\end{table}

\subsection{Equilibrium-temperature options}\label{sec:eqtemp}

Two options replace the full stochastic solve when an equilibrium approximation
is wanted:

\paragraph{(1) Equilibrium temperature for each grain type and size.}
Set \code{m\%stoch\_method = 'equil'} and call \code{dust\_emission}. Every grain
is placed at its own equilibrium temperature \Teq, computed from \emph{that
grain's} absorption cross section, and emits $C_{\rm abs}\,B_\lambda(\Teq)$. No
stochastic excursions are computed.

\paragraph{(2) A single equilibrium temperature for the whole model.}
\begin{lstlisting}
call dust_emission_single_teq(m, J_lam, lamI_total [, Teq_out])
\end{lstlisting}
All dust shares one temperature, found from the type- and size-integrated
(\,dn-weighted\,) total absorption cross section
$C_{\rm abs}^{\rm tot}(\lambda) = \sum_{\rm pop}\sum_a {\rm d}n(a)\,C_{\rm abs}(\lambda,a)$.
The single \Teq\ satisfies $\int C_{\rm abs}^{\rm tot}\,J\,{\rm d}\lambda =
\int C_{\rm abs}^{\rm tot}\,B(\Teq)\,{\rm d}\lambda$, and the emission is
$C_{\rm abs}^{\rm tot}\,B_\lambda(\Teq)$. The optional \code{Teq\_out} returns that
temperature.

Both options conserve energy (the emitted bolometric power equals the absorbed
power); they differ from the stochastic solve only in the SED \emph{shape} ---
a smooth modified blackbody rather than one carrying PAH/NIR features.

\subsection{Precomputed table + interpolation}

When the field in each cell is a fixed spectral \emph{shape} scaled by an intensity
$U$, precompute a table once and interpolate per cell:
\begin{lstlisting}
type(dust_emis_table_t) :: tab
call dust_build_table(m, J_ref, U_grid, tab [, status])            ! once
call dust_emission_interp(tab, U, lamI_total [, lamI_chan] [, status])  ! per cell
\end{lstlisting}
\code{J\_ref(m\%NLAM)} is the reference field shape (at $U{=}1$) and
\code{U\_grid(:)} an ascending intensity grid. The table is built from exact
solves, so it is exact at the grid points; interpolation is log--log in $U$.
The stored emissivities are floored at $10^{-300}$ before the log, so a grid
node whose true value is $0$ comes back as $10^{-300}$ rather than exactly $0$.
Both calls take an optional final \code{status} (\S\ref{sec:status}).

\paragraph{Caveat.} Stochastic emission is a \emph{nonlinear} functional of the
full $J(\lambda)$, not of a scalar. The table is valid only when the cell field
is $U\times(\text{fixed }J_{\rm ref})$. For cells whose field \emph{shape}
departs from $J_{\rm ref}$ (e.g.\ hardened spectra near hot stars), use the
cell-by-cell exact \code{dust\_emission}.

\subsection{Optional error status}\label{sec:status}

The builders and the emission/table calls each accept an optional final
\code{status} argument (\code{integer, intent(out)}). When it is present, an
invalid model, argument, or input file is reported through it (\code{status = 0}
on success, nonzero on failure) and the process keeps running, so a 3D RT host
can recover; when it is omitted, the same condition prints a message and
\code{stop}s the run, which the CLI drivers rely on. For the builders the
nonzero value only names the failing stage: the contract is simply
\code{0} = built, nonzero = build failed.

\begin{itemize}
\item \code{dust\_emission}: \code{1} unknown \code{stoch\_method};
      \code{2} \code{'qm'} selected but a population's radii are unset.
\item \code{dust\_build\_table}: \code{1} \code{size(J\_ref)}$\,\ne$\code{m\%NLAM};
      \code{2} \code{size(U\_grid)}$\,<2$;
      \code{3} \code{U\_grid} not positive and strictly increasing.
\item \code{dust\_emission\_interp}: \code{1} \code{U}$\,\le0$;
      \code{2} \code{size(lamI\_total)}$\,\ne$\code{tab\%NLAM};
      \code{3} \code{lamI\_chan} present but not
      \code{(tab\%NLAM, tab\%n\_channel)}.
\item \code{build\_astrodust}, \code{build\_dl07}: \code{1} $Q$-table load failed;
      \code{2} size-distribution load failed.
\item \code{build\_zubko}: \code{1} config read; \code{2} fewer than 3 components;
      \code{3} a component's optics read; \code{4} grid inconsistent;
      \code{5} a component's calorimetry read failed.
\item \code{build\_from\_files}: \code{1} descriptor open;
      \code{2} too many \code{pop:} lines; \code{3} invalid channel;
      \code{4} no \code{pop:} lines; \code{5} optics read;
      \code{6} grid inconsistent; \code{7} size-distribution read;
      \code{8} calorimetry read failed.
\item \code{dust\_extinction}: \code{1} an output array is not of size
      \code{m\%NLAM}.
\end{itemize}

%====================================================================
\section{Computing extinction}\label{sec:ext}

\code{dust\_extinction} is the extinction counterpart of \code{dust\_emission}:
a transfer code takes its opacity from the same model object, and on the same
wavelength grid \code{m\%lam}, as its emission.

\begin{lstlisting}
call dust_extinction(m, Cext, Cabs, Csca [, gbar] [, Cpol_ext] [, Cbir_ext] [, status])
\end{lstlisting}
\begin{itemize}
\item \code{Cext}, \code{Cabs}, \code{Csca} \code{(m\%NLAM)}: extinction,
      absorption and scattering cross sections per H atom [\,cm$^2$/H\,], with
      $C_{\rm ext} = C_{\rm abs} + C_{\rm sca}$.
\item \code{gbar(m\%NLAM)} (optional): the scattering-weighted asymmetry
      $\langle\cos\theta\rangle$, and $0$ where nothing scatters.
\item \code{Cpol\_ext(m\%NLAM)} (optional): the dichroic (polarized)
      extinction [\,cm$^2$/H\,]; see \S\ref{sec:pol}.
\item \code{Cbir\_ext(m\%NLAM)} (optional): the birefringent extinction
      [\,cm$^2$/H\,], the $f_{\rm align}$-weighted size integral of the
      orientation-resolved table's 4th block. It is $0$ when the loaded table
      carries no birefringence (the release table does not); see \S\ref{sec:pol}.
\item \code{status} (optional): error report; see \S\ref{sec:status}.
\end{itemize}

\noindent Each output is the plain binned sum over every population of the
model, since ${\rm d}n(a)$ already carries the bin width:
\begin{align}
C_{\rm abs}(\lambda) &= \sum_{\rm pop}\sum_a {\rm d}n(a)\,C_{\rm abs}(\lambda,a), &
C_{\rm sca}(\lambda) &= \sum_{\rm pop}\sum_a {\rm d}n(a)\,C_{\rm sca}(\lambda,a), \nonumber\\
\langle\cos\theta\rangle &= \frac{\sum {\rm d}n\,C_{\rm sca}\,g}{\sum {\rm d}n\,C_{\rm sca}}, &
C_{\rm pol,ext}(\lambda) &= \sum_{\rm pop}\sum_a {\rm d}n(a)\,C_{\rm pol,ext}(\lambda,a)\,f_{\rm align}(a).
\label{eq:extsums}
\end{align}
A population that carries no scattering or polarized optics contributes zero to
those terms. In the astrodust model the PAHs are exactly that case: they enter
through absorption only, and the astrodust grains carry all the scattering and
the asymmetry.

The values agree with the corresponding columns of
\code{data/kext\_astrodust\_MW.dat} (Table~\ref{tab:kextcols}) to the precision
with which that file is written (a median relative deviation of
$\sim7\times10^{-9}$ in each cross section), because both evaluate the same
size integral over the same optics. The difference is that the accessor returns
the result on \code{m\%lam} directly, so a host using its own wavelength grid
need not interpolate off the grid the driver happened to write.

\paragraph{Astrodust only for scattering.} \code{build\_dl07},
\code{build\_zubko} and \code{build\_from\_files} do not populate the
extinction-side optics, so for those models \code{Csca}, \code{gbar} and
\code{Cpol\_ext} come back as zeros while \code{Cabs} (and hence \code{Cext})
is still the correct absorption sum.

%====================================================================
\section{Polarized dust optics}\label{sec:pol}

The library exports two polarized quantities for the HD23 astrodust model: a
polarized \emph{emission} spectrum through \code{dust\_emission}, and a
polarized \emph{extinction} cross section through \code{dust\_extinction} or,
equivalently, through the precomputed extinction table. Both
are \emph{intrinsic}: the size integral and the alignment weighting are done,
but the geometry of the local magnetic field is not applied. What remains for
the calling code is set out in \S\ref{sec:pol:division}.

This section documents the interface only. For the alignment physics, the
polarized transfer equation, and the sign and index conventions behind these
quantities, see the companion report \code{aligned\_grain\_polarization.pdf}
(\S3 for the optical properties, \S4 for alignment, \S5 for transfer).

\subsection{Polarized emission}

\begin{lstlisting}
call dust_emission(m, J_lam, lamI_total [, lamI_chan] [, status] [, lamI_pol])
\end{lstlisting}
\begin{itemize}
\item \code{lamI\_pol(m\%NLAM)} (optional): the intrinsic polarized emission,
      with the same shape and the same units as \code{lamI\_total}. It is the
      emission a population of perfectly aligned grains would radiate with
      their symmetry axes in the plane of the sky.
\item Only populations that carry both polarized optics and an alignment
      efficiency contribute. PAHs are taken to be unaligned by HD23 and add
      zero, as does any model built without polarized optics.
\item The argument is optional and comes last, so every existing call site
      remains valid.
\end{itemize}

\noindent The polarized channel is accumulated at every point where the total
emission is accumulated, using the same grain temperature weights, so a
stochastically heated grain contributes to both consistently under all four
values of \code{m\%stoch\_method}.

\subsection{Polarized extinction}

Two routes give the same quantity. The direct one is the optional
\code{Cpol\_ext} argument of \code{dust\_extinction} (\S\ref{sec:ext}),
\begin{lstlisting}
call dust_extinction(m, Cext, Cabs, Csca, gbar, Cpol_ext)
\end{lstlisting}
which returns
$\sum_a {\rm d}n_{\rm Ad}(a)\,C_{\rm pol,ext}(\lambda,a)\,f_{\rm align}(a)$
[\,cm$^2$/H\,] on \code{m\%lam}. Prefer it in a code that already holds a model
object: it needs no file parsing and no interpolation onto a foreign grid.

The second route is the file \code{data/kext\_astrodust\_MW.dat}, written by
\code{calc\_kext\_astrodust.x}, which carries the dichroic extinction as an
eighth column (Table~\ref{tab:kextcols}). It remains the right choice for a
tool that wants the tabulated optics without linking the library. Codes that
read only the first seven columns are unaffected by its presence.

\begin{table}[h]
\centering
\begin{tabular}{@{}clll@{}}
\toprule
col. & quantity & unit & note \\
\midrule
1 & \code{lambda}        & \um\        & wavelength grid \\
2 & \code{albedo}        &             & $C_{\rm sca}/C_{\rm ext}$ \\
3 & \code{<cos>}         &             & scattering asymmetry $g$ \\
4 & \code{C\_ext/H}      & cm$^2$/H    & total extinction \\
5 & \code{C\_abs/H}      & cm$^2$/H    & total absorption \\
6 & \code{C\_sca/H}      & cm$^2$/H    & total scattering \\
7 & \code{K\_abs}        & cm$^2$/g    & mass absorption coefficient \\
8 & \code{C\_polext/H}   & cm$^2$/H    & dichroic extinction (new) \\
\bottomrule
\end{tabular}
\caption{Columns of \code{kext\_astrodust\_MW.dat}. Column 8 is
$\sum_a {\rm d}n_{\rm Ad}(a)\,C_{\rm pol,ext}(\lambda,a)\,f_{\rm align}(a)$,
the maximum polarized extinction, quoted before any geometric factor.}
\label{tab:kextcols}
\end{table}

\subsection{Changing the alignment efficiency}\label{sec:pol:align}

By default the alignment efficiency is the HD23 fit. Two setters, both
re-exported by \code{dust\_lib}, replace it on a model that already exists.

\begin{lstlisting}
call dust_set_alignment(m, f_max, a_align, alpha_align [, status])
call dust_set_alignment_profile(m, aeff_in, falign_in [, status])
\end{lstlisting}

The first installs the HD23 power-law rolloff
\begin{equation}
f_{\rm align}(a) \;=\; \frac{f_{\max}}{1 + (a_{\rm align}/a)^{\alpha_{\rm align}}},
\label{eq:falignset}
\end{equation}
with $a_{\rm align}$ in \um. The second installs an arbitrary tabulated
profile: the caller supplies $(\code{aeff\_in}\,[\um],\ \code{falign\_in})$
pairs and each population's efficiency is interpolated from them onto its own
radius grid, linearly in $\log a$ and clamped at the ends of the table. This
is the route for prescriptions Eq.~(\ref{eq:falignset}) cannot express: a
RAT-derived Rayleigh reduction factor $R(a)$, or the GRADE-POL exponential
$f_{\max}[1 - \exp(-(0.5a/a_{\rm align})^3)]$.

Both refill \code{falign} only for populations that carry polarized optics.
PAHs, which HD23 take to be unaligned, are left untouched and keep
contributing zero.

\paragraph{No temperature solve is repeated.} This is the property worth
knowing before designing a parameter study. The alignment efficiency enters as
a weight in the size integral, \emph{outside} the temperature solution: it
multiplies the polarized accumulator and never enters $P(T)$. Changing the
alignment therefore does not invalidate any temperature distribution, and
\code{lamI\_total} is bit-for-bit unchanged across a setter call. A scan over
alignment parameters costs one \code{dust\_emission} call per point, not one
model build.

\paragraph{Negative efficiencies are allowed on purpose.}
\code{dust\_set\_alignment\_profile} requires \code{falign\_in} to lie in
$[-1, 1]$ and \emph{rejects} rather than clamps a value outside it, since
$|f| \le 1$ is the physical bound on a Rayleigh reduction factor and a value
beyond it is a caller error that silent clamping would hide. Negative values
inside the range are accepted: a grain with slow internal relaxation has
$Q_X = -0.1$, a negative reduction factor that flips the polarization
direction by $90^\circ$. That is the accepted origin of polarization parallel
to $\mathbf{B}$ at millimeter wavelengths and perpendicular to it in the
submillimeter, and clamping at zero would silently delete it.
\code{dust\_set\_alignment} keeps $f_{\max}$ in $[0, 1]$, the power law having
no such sign freedom.

With \code{status} present a bad argument sets it nonzero and returns; without
it, the run stops. The codes are 1, 2, 3 in the order the arguments are listed
above for \code{dust\_set\_alignment} ($a_{\rm align} \le 0$,
$\alpha_{\rm align} \le 0$, $f_{\max}$ outside $[0,1]$), and for
\code{dust\_set\_alignment\_profile} 1 for mismatched or too-short arrays,
2 for an \code{aeff\_in} that is not positive and strictly increasing, and
3 for a \code{falign\_in} value outside $[-1, 1]$.

\paragraph{Status code 4: a loaded aligned scattering table is now stale.}
If an aligned scattering table has been loaded (\S\ref{sec:pol:scatmat}) and the
requested profile differs from the one that table was integrated under, either
setter returns the new code \code{4}. This is \emph{non-fatal}: the
$f_{\rm align}$ update completes and the emission and \code{Cpol\_ext} channels
re-integrate live under the new profile, so those outputs always honor the latest
alignment. The code only signals that the loaded scattering matrices no longer
correspond to it, because they were size-integrated once under the recorded
profile. The sanctioned runtime variation of the aligned scattering optics is the
scalar $\eta$ (\S\ref{sec:pol:scatmat}); a genuinely different profile requires
regenerating the table with \code{run\_scatmat\_aligned.x profile=FILE}. A
tabulated profile never matches the recorded analytic one and so always raises
the flag.

\subsection{Where the polarized optics come from}\label{sec:pol:source}

The orientation-resolved spheroid cross sections behind \code{lamI\_pol} and
\code{Cpol\_ext} default to the Draine \& Hensley (2021) release table. They can
instead be taken from a table SEDust regenerates from the astrodust dielectric
function with its own T-matrix engine: \code{build\_astrodust} and \code{sed\_init}
accept an optional \code{qpol\_path} (with \code{qpol\_wave\_path} and
\code{qpol\_aeff\_path} for its grid axes) that overrides the default file. The
regenerated table matches the release to a few parts in $10^4$ where grains carry
weight and is written in the release table's own format, so nothing else in the
call changes. Its
one deliberate benefit is that it supplies the polarized cross sections from the
same exact optics as the total intensity, removing the mixed averaging convention
the release forces; see \code{sedust\_polarization\_implementation.pdf} \S7 and
\code{tmatrix/run\_q\_jori.x}. Leaving \code{qpol\_path} unset keeps the release
table and the existing results.

\subsection{Aligned-grain scattering optics}\label{sec:pol:scatmat}

For the astrodust spheroid SEDust also supplies the angle-resolved scattering
optics of the aligned population, for a polarized RT code that transports a Stokes
vector: the fixed-orientation Mueller matrix, the $4\times4$ extinction matrix,
and the random-orientation remainder. These come from a table written by
\code{tmatrix/run\_scatmat\_aligned.x} (five optical bands ship with SEDust); how
it is built is documented in \code{sedust\_polarization\_implementation.pdf}
\S8. The API follows the library's own lifecycle --- initialize once, then query
along the photon path --- split into two strictly separated layers.

\paragraph{Loading (once per run, serial).} Pass the table path as an optional
\code{scatmat\_path} to \code{sed\_init} or \code{build\_astrodust}, exactly like
\code{qpol\_path}:
\begin{lstlisting}
call build_astrodust(m, qtab, sizedist, NT, T_lo, T_hi, status=st, &
                     qpol_path=QPOL, scatmat_path=SCATMAT)
\end{lstlisting}
The table is parsed and its size integrals kept in memory (about $81.5$\,MB for the
production grid). A \code{scatmat\_path} that fails to load is an error
(\code{status = 3} from the builder), because a host that asked for the table
depends on it. This layer is not thread-safe; call it from serial code before the
photon loop.

\paragraph{Path queries (per event, thread-safe).} Five routines, all re-exported
by \code{dust\_lib}, are pure reads of the loaded arrays: they write only their own
output arguments, do no I/O and no allocation, and are safe to call concurrently
from OpenMP photon threads. The whole cell dependence enters through two scalars
the host already holds --- the alignment scale $\eta$, and the incidence angle
$\theta_i = \arccos(\hat k\!\cdot\!\hat B)$ between the photon direction and the
local field, from one dot product.
\begin{lstlisting}
call scatmat_band(lambda_um, iband, exact)                     ! pick the band once
call extinction_matrix_aligned(iband, theta_i, eta, kmat)      ! 4x4  [um^2/H]
call mueller_matrix_aligned(iband, theta_i, theta_s, phi, z)   ! 4x4  [um^2 sr^-1/H], eta=1
call mueller_matrix_random(iband, big_theta, f_tot, f_ref)     ! 6-elt random matrices
call scattering_cross_sections(iband, theta_i, eta, csca_aligned, csca_unaligned)
\end{lstlisting}
\begin{itemize}
\item \code{scatmat\_band}: \code{lambda\_um} [\um] in; \code{iband} the nearest
      stored band; \code{exact} \code{.true.} if it matched to $10^{-3}$
      fractionally. The hot path then takes \code{iband} so no wavelength search
      runs each event.
\item \code{extinction\_matrix\_aligned}: \code{kmat(4,4)} [\,\um$^2$/H\,], the
      $\eta$-scaled aligned extinction matrix at incidence \code{theta\_i} [deg].
      Diagonal $\eta\,C_{\rm ext,al}$; $K(1,2)=K(2,1)=\eta\,C_{\rm pol,al}$
      (dichroism); $K(3,4)=\eta\,C_{\rm bir,al}$, $K(4,3)=-\eta\,C_{\rm bir,al}$
      (birefringence). \code{theta\_i} in $[0,180^\circ]$ is folded to
      $[0,90^\circ]$, $K$ being even under the equatorial reflection.
\item \code{mueller\_matrix\_aligned}: \code{z(4,4)} [\,\um$^2$\,sr$^{-1}$/H\,] at
      the reference alignment ($\eta=1$; scale it yourself). \code{theta\_i},
      \code{theta\_s} in $[0,180^\circ]$, \code{phi} in $[0,360^\circ]$; the
      stored octants ($\theta_i\!\le\!90^\circ$, $\phi\!\le\!180^\circ$) and the
      rest are handled by interpolation plus the header's symmetry relations.
\item \code{mueller\_matrix\_random}: \code{f\_tot(6)}, \code{f\_ref(6)}, the two
      six-element ($F_{11},F_{22},F_{33},F_{44},F_{12},F_{34}$) random-orientation
      matrices at \code{big\_theta} [deg], $\alpha_1$-normalized as stored.
      Restore absolute \um$^2$\,sr$^{-1}$/H with \code{f\_tot*scm\_csca\_tot(iband)}
      and \code{f\_ref*scm\_csca\_ref(iband)}.
\item \code{scattering\_cross\_sections}: \code{csca\_aligned} $=\eta\,
      C_{\rm sca,al}(\theta_i)$ and \code{csca\_unaligned}
      $=C_{\rm sca,tot}-\eta\,C_{\rm sca,ref}$ [\,\um$^2$/H\,], for the albedo and
      the population split.
\end{itemize}

\paragraph{The $\eta$ contract, and unit note.} A cell's alignment enters only as
$\eta$: the aligned matrix and the aligned $K$ scale as $\eta$, and the unaligned
remainder scattering matrix is $F_{\rm tot}-\eta F_{\rm ref}$ in absolute units.
For the total extinction matrix the aligned population must not be counted twice:
the isotropic \code{Cext} from \code{dust\_extinction} already contains its
reference-orientation average $C_{\rm ext,ref}$, so assemble
\begin{equation}
  K_{\rm total} = (C_{\rm ext}^{\rm iso} - \eta\,C_{\rm ext,ref})\,I
                  + \eta\,K_{\rm al}(\theta_i),
\label{eq:ktotal}
\end{equation}
which returns $C_{\rm ext}^{\rm iso}$ on the diagonal at the incidence average and
adds the direction-resolved dichroism and birefringence. These aligned-scattering
quantities are in \um$^2$ (per H), whereas \code{dust\_extinction} returns cm$^2$/H;
convert with a factor $10^{8}$ before combining them, as
Eq.~(\ref{eq:ktotal}) requires.

\paragraph{Inlining.} The loaded grids and integrals are also exposed
\code{public, protected} (\code{scm\_lambda}, \code{scm\_theta\_i},
\code{scm\_theta\_s}, \code{scm\_phi}, \code{scm\_theta\_ran}, \code{scm\_cext\_tot},
\code{scm\_csca\_tot}, \code{scm\_cext\_ref}, \code{scm\_csca\_ref},
\code{scm\_F\_tot}, \code{scm\_F\_ref}, \code{scm\_Z}, and the \code{scm\_n*}
sizes), so a host that prefers to inline its own lookups can take them once at
startup and never call back. A complete two-cell reference consumer of every call
is \code{sed/rt\_example/use\_dustlib\_scatmat.f90} (\code{make
use\_dustlib\_scatmat.x}); it demonstrates the band selection,
Eq.~(\ref{eq:ktotal}), the aligned and random queries, and the closure checks.

\subsection{Division of labor}\label{sec:pol:division}

Table~\ref{tab:poldivision} is the part to get right. SEDust stops at the
intrinsic quantity; everything that depends on where the field points is left
to the transfer code, because SEDust knows nothing about the cell geometry.

\begin{table}[h]
\centering
\begin{tabular}{@{}p{0.46\textwidth}p{0.46\textwidth}@{}}
\toprule
Done by SEDust & Left to the calling code \\
\midrule
Integral over the grain size distribution &
$\sin^2\gamma$, with $\gamma$ the angle between the \emph{local magnetic
field} and the \emph{line of sight} (not a sky-plane angle) \\[2pt]
Weighting by the alignment efficiency $f_{\rm align}(a)$ &
Turbulent depolarization $F_{\rm turb}$ \\[2pt]
Summation over components (only astrodust is aligned; PAHs contribute zero) &
Splitting into $Q$ and $U$ using the position angle $\phi$ of the projected
field \\
\bottomrule
\end{tabular}
\caption{What is already integrated into \code{lamI\_pol} and into
\code{Cpol\_ext} (equivalently, column 8), and what a radiative transfer host
must still apply.}
\label{tab:poldivision}
\end{table}

\noindent The observed Stokes emissivities follow from the exported quantity as
\begin{equation}
(j_Q,\,j_U) \;=\; \code{lamI\_pol}\,\sin^2\!\gamma\;F_{\rm turb}\,
                  (\cos 2\phi,\;\sin 2\phi),
\label{eq:stokesuse}
\end{equation}
with $j_I = \code{lamI\_total}$ unchanged. The same $\sin^2\gamma$ and
$F_{\rm turb}$ multiply \code{Cpol\_ext} (column 8) when the extinction matrix
is assembled.
Equation~(\ref{eq:stokesuse}) is stated here as a usage rule; its derivation is
in \S5.1 of \code{aligned\_grain\_polarization.pdf}.

\subsection{Minimal example}

\begin{lstlisting}
use dust_lib
use radfield, only: J_Mathis
type(dust_model_t)    :: m
real(wp), allocatable :: J_lam(:), lamI_total(:), lamI_pol(:), p(:)
integer :: i

call build_astrodust(m, qtab, sizedist, 200, 2.7_wp, 5.0e3_wp)
allocate(J_lam(m%NLAM), lamI_total(m%NLAM), lamI_pol(m%NLAM), p(m%NLAM))

call J_Mathis(1.0_wp, m%lam, J_lam)
call dust_emission(m, J_lam, lamI_total, lamI_pol=lamI_pol)

! intrinsic polarization fraction (before any geometric factor)
do i = 1, m%NLAM
   if (lamI_total(i) > 0.0_wp) then
      p(i) = lamI_pol(i) / lamI_total(i)
   else
      p(i) = 0.0_wp
   end if
end do
\end{lstlisting}

\noindent Multiply \code{p} by $\sin^2\gamma\,F_{\rm turb}$ to obtain the
polarization fraction of a real line of sight.

\subsection{Limits}

\begin{itemize}
\item \textbf{Astrodust only.} \code{build\_dl07} and \code{build\_zubko} have
      no polarized optics. \code{build\_dl07} releases any polarized arrays left
      by a previous astrodust build, so \code{lamI\_pol} returns zeros rather
      than stale values.
\item \textbf{PAHs contribute zero} ($f_{\rm align}=0$, an HD23 assumption).
      The polarized emission therefore vanishes in the mid infrared, where
      stochastically heated PAHs dominate.
\item \textbf{Circular polarization: available only from the regenerated table.}
      The released optics table carries $Q_{\rm ext}$, $Q_{\rm abs}$ and
      $Q_{\rm sca}$ but no phase retardation, so with it $C_{\rm circ}=0$. The
      regenerated orientation-resolved table (\S\ref{sec:pol:source}) adds a
      birefringence block, and \code{dust\_extinction} returns it as the optional
      \code{Cbir\_ext} (\S\ref{sec:ext}); the aligned scattering table carries the
      same birefringence in its $K(\theta_i)$ (\S\ref{sec:pol:scatmat}).
      Consuming it in the $U\leftrightarrow V$ transfer term is the host's task.
\item \textbf{Scattering by aligned grains: now available.} The angle-resolved
      Mueller matrix of the aligned astrodust spheroid, which no released file
      contains, is computed by \code{run\_scatmat\_aligned.x} and served through
      the API of \S\ref{sec:pol:scatmat}. The release-derived cross sections still
      supply only the integrated $Q_{\rm sca}$; the angular optics come from the
      T-matrix engine.
\item \textbf{Far-infrared and submillimeter polarized emission is complete
      without a scattering matrix.} The albedo there is essentially zero, so the
      transfer needs only the extinction matrix and the emission vector, both
      fully determined by what is exported here.
\end{itemize}

\paragraph{Reference numbers.} For a Mathis ISRF the intrinsic polarization
fraction \code{lamI\_pol/lamI\_total} is $0.00\%$ at 12\um, $13.67\%$ at
100\um, $17.15\%$ at 154\um, $18.79\%$ at 350\um\ and $19.15\%$ at 850\um.
Column 8 reproduces the released \code{polarized\_extinction.dat} to a median
of $0.0296\%$.

%====================================================================
\section{The model object and accessors}

\begin{lstlisting}
type :: dust_model_t
   character(len=32)              :: name
   integer                       :: NA, NLAM, NT
   real(wp), allocatable         :: lam(:)        ! (NLAM) [um]
   real(wp), allocatable         :: T_first(:)    ! (NT)   [K]
   type(grain_pop_t), allocatable :: pops(:)
   integer                       :: n_channel
   character(len=16), allocatable :: channel_name(:)
   logical                       :: use_induced_emission   ! default .false.
   character(len=16)             :: stoch_method           ! default 'heuristic'
   logical                       :: verbose                ! default .false.
end type
\end{lstlisting}

\paragraph{Diagnostics.} \code{m\%verbose} (default \code{.false.}) keeps the
library solve path silent; set it \code{.true.} to print the same solver
diagnostics as the CLI drivers.

Convenience accessors (all in \code{dust\_lib}):
\begin{lstlisting}
n  = dust_nlam(m)              ! number of wavelengths
nc = dust_n_channel(m)         ! number of output channels
lam = dust_lambda(m)           ! allocatable copy of the wavelength grid [um]
nm  = dust_channel_name(m, ic) ! name of channel ic
\end{lstlisting}

%====================================================================
\section{The generic file loader}

\code{build\_from\_files} reads a small text \emph{descriptor} and assembles a
model from data files, with no code changes. One \code{pop:} line per
population:
\begin{lstlisting}[style=sh]
# descriptor.txt
name = mymodel
# pop: <grain_type> <channel> <optics> <dnda_table> <calorimetry> <rho(0=use file)>
pop: pah 1 PAH_28_1201_neu.dat  SzDist_PAH.dat Graphitic_Calorimetry.dat 0.0
pop: gra 2 Gra_121_1201.dat     SzDist_Gra.dat Graphitic_Calorimetry.dat 0.0
pop: sil 3 suvSil_121_1201.dat  SzDist_Sil.dat Silicate_Calorimetry.dat  0.0
\end{lstlisting}
\begin{itemize}
\item \code{grain\_type}: \code{'sil'}, \code{'pah'}, or \code{'gra'} (selects the
      heat-capacity/mode model used by the \code{'qm'} solver; ignored by the GD
      solvers, which use the supplied enthalpy directly).
\item \code{channel}: integer output channel ($1\dots$); populations sharing a
      channel are summed.
\item \code{optics}: a DustEM/Zubko $Q$-table ($Q_{\rm abs}$ per radius and
      wavelength); $C_{\rm abs}=Q_{\rm abs}\pi a^2$.
\item \code{dnda\_table}: a 2-column table $a$ [\um] and ${\rm d}n/{\rm d}a$
      [cm$^{-1}$ H$^{-1}$].
\item \code{calorimetry}: a specific-enthalpy table $u(T)$ [erg/g];
      $H(T,a)=u(T)\,\rho\,\tfrac{4\pi}{3}a^3$.
\item \code{rho}: grain density [g\,cm$^{-3}$]; \code{0.0} uses the value in the
      optics-file header.
\end{itemize}
All files are sought under \code{data\_dir}. The coded builders
(\code{build\_astrodust}/\code{dl07}/\code{zubko}) are the recommended path for
the built-in models; \code{build\_from\_files} is for adding new data-defined
models.

%====================================================================
\section{Linking into a 3D RT code}

Compile your RT source against \code{libsedust.a} with the \code{.mod} search path
pointing at \code{SEDust/sed/}:
\begin{lstlisting}[style=sh]
gfortran -I<sed> my_rt.f90 <sed>/libsedust.a -fopenmp -o my_rt.x
\end{lstlisting}
A complete minimal example is provided in
\code{SEDust/sed/rt\_example/use\_dustlib.f90}: it builds a model, computes
emission for one field, and prints the SED peak. No \code{ISO\_C\_BINDING} is
needed --- the RT code simply \code{use}s the Fortran module \code{dust\_lib}.

\paragraph{Threading.} \code{m} is read-only during \code{dust\_emission}; all
scratch is local. The solver uses OpenMP internally over grains. If your RT
parallelizes over cells, take care to avoid oversubscription (nested
parallelism); the default \code{'heuristic'} solver is serial per call, which
suits an outer cell-level parallel loop.

%====================================================================
\section{Units and conventions}

\begin{itemize}
\item Wavelengths \code{lam} are in \um\ throughout; cross sections in cm$^2$.
\item \code{J\_lam} is a mean intensity with the same units as the Planck
      function $B_\lambda$ (specific intensity per steradian). The library's
      output \lamIlam$/\Nh$ is then in
      erg\,s$^{-1}$\,cm$^{-2}$\,sr$^{-1}$\,H$^{-1}$.
\item The CMB (2.725\,K) is included in the heating field by \code{J\_Mathis}.
\item The induced-/stimulated-emission factor is off by default (the output is
      \emph{net} emission, matching the HD23 and DL07spec released SEDs).
\end{itemize}

%====================================================================
\section{Data files}

Model data lives under \code{SEDust/data/}:
\begin{itemize}
\item astrodust / DL07: the T-matrix $Q$-table
      (\code{../tmatrix/output/q\_astrodust\_*.dat}) and the HD23 release size
      distribution (\code{../data/release/size\_distribution.dat}).
\item Zubko: \code{../data/zubko/} --- the ZDA config/formula, the tabulated
      size distributions, the DustEM $Q$-tables, and the calorimetry tables
      from the Camps et al.\ 2015 benchmark; a ready descriptor is
      \code{../data/zubko/zubko\_descriptor.txt}.
\end{itemize}

\end{document}
